---
title: "Statistical Models"
subtitle: "Lecture 4"
from: markdown+emoji
author: 
    - name: Dr. Silvio Fanzon
      id: sf
      email: S.Fanzon@hull.ac.uk
      url: https://www.silviofanzon.com
      affiliations: University of Hull
---



::: {.content-hidden}
$
{{< include macros.tex >}}
$
:::




# Lecture 4: <br> Two-sample t-test & <br> More on R  {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Outline of Lecture 4


::: {.column width="57%"}

1. Two-sample hypothesis tests
2. Two-sample t-test
3. Two-sample t-test: Example
4. The Welch t-test
5. The t-test for paired samples

:::



::: {.column width="42%"}

6. More on vectors

:::







# Part 1: <br>Two-sample<br> hypothesis tests {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Overview {.smaller}

In Lecture 3:

- We looked at data on CCI before and after the 2008 crash
- In this case data for each month is directly comparable
- Can then construct the difference between the 2007 and 2009 values
- Analysis reduces from a two-sample to a one-sample problem

::: Question

How do we analyze two samples that cannot be paired?

:::





## Problem statement {.smaller}

**Goal:** compare mean and variance of 2 independent **normal** samples

- First sample:
    * $X_1, \ldots, X_n$ from normal population $N(\mu_X,\sigma_X^2)$

- Second sample:
    * $Y_1, \ldots, Y_m$ from normal population $N(\mu_Y,\sigma_Y^2)$

- We may have $n \neq m$ 
    * Samples cannot be paired due to different size!

**Tests available:**

- Two-sample $t$-test to test for difference in means
- Two-sample $F$-test to test for difference in variances (next week)




## Why is this important? {.smaller}

- Hypothesis testing starts to get interesting with 2 or more samples

- t-test and F-test show the normal distribution family in action

- This is also the maths behind regression
    * Same methods apply to seemingly unrelated problems
    * Regression is a big subject in statistics




## Normal distribution family in action {.smaller}
### Two-sample t-test

- Want to compare the means of two independent samples
- At the same time population variances are unknown
- Therefore both variances are estimated with sample variances
- Test statistic is $t_k$-distributed with $k$ linked to the total number of observations




## Normal distribution family in action {.smaller}
### Two-sample F-test

- Want to compare the variance of two independent samples
- This can be done by studying the ratio of the sample variances
$$
S^2_X/S^2_Y
$$

- We have already shown that
$$
\frac{(n - 1) S^2_X}{\sigma^2_X} \sim \chi^2_{n - 1} \qquad 
\frac{(m - 1) S^2_Y}{\sigma^2_Y} \sim \chi^2_{m - 1}
$$



## Normal distribution family in action {.smaller}
### Two-sample F-test


- Hence we can study statistic
$$
F = \frac{S^2_X / \sigma_X^2}{S^2_Y / \sigma_Y^2}
$$

- We will see that $F$ has F-distribution (next week)






# Part 2: <br>Two-sample t-test {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## The two-sample t-test {.smaller}

**Assumptions**: Suppose given samples from 2 **independent** normal populations

- $X_1, \ldots ,X_n$ iid with distribution $N(\mu_X,\sigma_X^2)$
- $Y_1, \ldots ,Y_m$ iid with distribution $N(\mu_Y,\sigma_Y^2)$


**Further assumptions**:

- In general $n \neq m$, so that one-sample t-test cannot be applied
- The two populations have same variance 
$$
\sigma^2_X = \sigma^2_Y = \sigma^2
$$


**Note:** Assuming same variance is simplification. Removing it leads to *Welch t-test*





## The two-sample t-test {.smaller}


**Goal**: Compare means $\mu_X$ and $\mu_Y$

**Hypothesis set:** We test for a difference in means
$$
H_0 \colon \mu_X = \mu_Y \qquad  H_1 \colon \mu_X \neq \mu_Y
$$


**t-statistic**: The general form is
$$
T = \frac{\text{Estimate}-\text{Hypothesised value}}{\text{e.s.e.}}
$$



## The two-sample t-statistic {.smaller}

- Define the sample means
$$
\overline{X} = \frac{1}{n} \sum_{i=1}^n X_i
\qquad \qquad
\overline{Y} = \frac{1}{m} \sum_{i=1}^m Y_i
$$

- Notice that
$$
\Expect[ \overline{X} ] = \mu_X   \qquad \qquad
\Expect[ \overline{Y} ] = \mu_Y
$$

- Therefore we can estimate $\mu_X - \mu_Y$ with the sample means, that is,
$$
\text{Estimate} = \overline{X} - \overline{Y}
$$




## The two-sample t-statistic {.smaller}

- Since we are testing for difference in mean, we have
$$
\text{Hypothesised value} = \mu_X - \mu_Y
$$

- The *Estimated Standard Error* is the standard deviation of estimator
$$
\text{e.s.e.} = \text{Standard Deviation of } \overline{X} -\overline{Y} 
$$


## The two-sample t-statistic {.smaller}

- Therefore the two-sample t-statistic is
$$
T = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{\text{e.s.e.}}
$$

- Under the Null Hypothesis that $\mu_X = \mu_Y$, the t-statistic becomes
$$
T = \frac{\overline{X} - \overline{Y} }{\text{e.s.e.}}
$$




## A note on the degrees of freedom (df) {.smaller}

- The general rule is
$$
\text{df} = \text{Sample size} - \text{No. of estimated parameters}
$$

- Sample size in two-sample t-test:
    * $n$ in the first sample
    * $m$ in the second sample 
    * Hence total number of observations is $n + m$

- No. of estimated parameters is 2: Namely $\mu_X$ and $\mu_Y$

- Hence degree of freedoms in two-sample t-test is
$$
{\rm df} = n + m - 2
$$
(more on this later)





## The estimated standard error {.smaller}

- Recall: We are assuming populations have same variance
$$
\sigma^2_X = \sigma^2_Y = \sigma^2
$$

- We need to compute the estimated standard error
$$
\text{e.s.e.} = \text{Standard Deviation of } \ \overline{X} -\overline{Y} 
$$

- Variance of sample mean was computed in the Lemma in Slide 72 Lecture 2 

-  Since $\overline{X} \sim N(\mu_X,\sigma^2)$ and $\overline{Y} \sim N(\mu_Y,\sigma^2)$, by the Lemma we get
$$
\Var[\overline{X}] = \frac{\sigma^2}{n} \,, \qquad \quad
\Var[\overline{Y}] = \frac{\sigma^2}{m}
$$



## The estimated standard error {.smaller}

- Since $X_i$ and $Y_i$ are independent we get $\Cov(X_i,Y_j)=0$

- By bilinearity of covariance we infer
$$
\Cov ( \overline{X} , \overline{Y} ) = \frac{1}{n \cdot m} \sum_{i=1}^n \sum_{j=1}^m \Cov (X_i,Y_j) = 0
$$

- We can then compute
\begin{align*}
\Var[ \overline{X} - \overline{Y} ] & = \Var[ \overline{X} ] + \Var [ \overline{Y} ] -
                                        2 \Cov( \overline{X} , \overline{Y} ) \\
                                    & = \Var[ \overline{X} ] + \Var [ \overline{Y} ] \\
                                    & = \sigma^2 \left( \frac{1}{n} + \frac{1}{m}  \right)
\end{align*}





## The estimated standard error {.smaller}

- Taking the square root gives
$$
\text{S.D.}(\overline{X} - \overline{Y} )= \sigma \ \sqrt{\frac{1}{n}+\frac{1}{m}}
$$

- Therefore, the t-statistic is
$$
T = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{\text{e.s.e.}} 
  = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{\sigma \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}}
$$







## Estimating the variance {.smaller}

The t-statistic is currently
$$
T = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{\sigma \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}}
$$

- Variance $\sigma^2$ is unknown: we need to **estimate it**!

- Define the sample variances

$$
S_X^2 = \frac{ \sum_{i=1}^n X_i^2 - n \overline{X}^2 }{n-1} \qquad \qquad
S_Y^2 = \frac{ \sum_{i=1}^m Y_i^2 - m \overline{Y}^2 }{m-1}
$$



## Estimating the variance {.smaller}

- Recall that 
$$
X_1, \ldots , X_n \sim N(\mu_X, \sigma^2) \qquad \qquad
Y_1, \ldots , Y_m \sim N(\mu_Y, \sigma^2)
$$

- From Lecture 2, we know that $S_X^2$ and $S_Y^2$ are unbiased estimators of $\sigma^2$, i.e.
$$
\Expect[ S_X^2 ] = \Expect[ S_Y^2 ] = \sigma^2
$$

- Therefore, both $S_X^2$ and $S_Y^2$ can be used to estimate $\sigma^2$



## Estimating the variance {.smaller}

- We can improve the estimate of $\sigma^2$ by combining  $S_X^2$ and $S_Y^2$

- We will consider a (convex) linear combination
$$
S^2 := \lambda_X S_X^2 + \lambda_Y S_Y^2  \,, \qquad  
\lambda_X + \lambda_Y = 1
$$


- $S^2$ is still an unbiased estimator of $\sigma^2$, since
\begin{align*}
\Expect[S^2]  & = \Expect[ \lambda_X S_X^2 + \lambda_Y S_Y^2 ] \\
              & = \lambda_X \Expect[S_X^2] + \lambda_Y \Expect[S_Y^2] \\
             & = (\lambda_X + \lambda_Y) \sigma^2 \\
             & = \sigma^2
\end{align*}



## Estimating the variance {.smaller}


We choose coefficients $\lambda_X$ and $\lambda_Y$ which reflect sample sizes
$$
\lambda_X := \frac{n - 1}{n + m - 2} \qquad 
\qquad
\lambda_Y := \frac{m - 1}{n + m - 2}
$$

**Notes**: 

- We have $\lambda_X + \lambda_Y = 1$

- Denominators in $\lambda_X$ and $\lambda_Y$ are degrees of freedom 
$${\rm df } = n + m - 2$$

- This choice is made so that $S^2$ has chi-squared distribution (more on this later) 




## Pooled estimator of variance {.smaller}

::: Definition

The **pooled estimator** of $\sigma^2$ is defined as
$$
S_p^2 := \lambda_X S_X^2 + \lambda_Y S_Y^2 
      = \frac{(n-1) S_X^2 + (m-1) S_Y^2}{n + m - 2}
$$

:::


**Note**: 

- $n=m$ implies $\lambda_X = \lambda_Y$
- In this case $S_X^2$ and $S_Y^2$ have same weight in $S_p^2$




## The two-sample t-statistic {.smaller}

- The t-statistic has currently the form
$$
T = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{\sigma \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}}
$$

- We replace $\sigma$ with the pooled estimator $S_p$


## The two-sample t-statistic {.smaller}

::: Definition

The two sample t-statistic is defined as
$$
T := \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{ S_p \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}}  
$$
:::

**Note**: Under the Null Hypothesis that $\mu_X = \mu_Y$ this becomes
$$
T = \frac{\overline{X} - \overline{Y}}{ S_p \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}} 
= \frac{\overline{X} - \overline{Y}}{ \sqrt{ \dfrac{ (n-1) S_X^2 + (m-1) S_Y^2 }{n + m - 2} } \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}} 
$$



## Distribution of two-sample t-statistic {.smaller}

::: Theorem 

The two sample t-statistic has $t_{n+m-2}$ distribution
$$
T := \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{ S_p \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}}  \sim t_{n + m - 2}
$$

:::



## Distribution of two-sample t-statistic {.smaller}
### Proof

- We have already seen that $\overline{X} - \overline{Y}$ is normal with
$$
\Expect[\overline{X} - \overline{Y}] = \mu_X - \mu_Y 
\qquad \qquad
\Var[\overline{X} - \overline{Y}] = \sigma^2 \left( \frac{1}{n} + \frac{1}{m} \right)
$$

- Therefore we can rescale $\overline{X} - \overline{Y}$ to get
$$
U := \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{ \sigma \sqrt{ \dfrac{1}{n} + \dfrac{1}{m}}} \sim N(0,1)
$$



## Distribution of two-sample t-statistic {.smaller}
### Proof


- We are assuming $X_1, \ldots, X_n$ iid $N(\mu_X,\sigma^2)$

- Therefore, as already shown, we have 
$$
\frac{ (n-1) S_X^2 }{ \sigma^2 } \sim \chi_{n-1}^2
$$

- Similarly, since $Y_1, \ldots, Y_m$ iid $N(\mu_Y,\sigma^2)$, we get
$$
\frac{ (m-1) S_Y^2 }{ \sigma^2 } \sim \chi_{m-1}^2
$$




## Distribution of two-sample t-statistic {.smaller}
### Proof


- Since $X_i$ and $Y_j$ are independent, we also have that
$$
\frac{ (n-1) S_X^2 }{ \sigma^2 }   \quad \text{ and } \quad 
\frac{ (m-1) S_Y^2 }{ \sigma^2 } \quad \text{ are independent}
$$


- In particular we obtain
$$
\frac{ (n-1) S_X^2 }{ \sigma^2 }  + \frac{ (m-1) S_Y^2 }{ \sigma^2 } \sim \chi_{n-1}^2 + \chi_{m-1}^2 \sim \chi_{m + n- 2}^2
$$



## Distribution of two-sample t-statistic {.smaller}
### Proof

- Recall the definition of $S_p^2$
$$
S_p^2 = \frac{(n-1) S_X^2 + (m-1) S_Y^2}{ n + m - 2 }
$$

- Therefore
$$
V := \frac{ (n+m-2) S_p^2 }{ \sigma^2 } 
= \frac{ (n - 1) S_X^2}{ \sigma^2}  + \frac{ (m-1) S_Y^2 }{ \sigma^2 } \sim \chi_{n + m - 2}^2 
$$



## Distribution of two-sample t-statistic {.smaller}
### Proof

- Rewrite $T$ as
\begin{align*}
T & = \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{ S_p \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}} \\
  & =  \frac{\overline{X} - \overline{Y} - (\mu_X - \mu_Y)}{ \sigma \sqrt{ \dfrac{1}{n} + \dfrac{1}{m} }   } 
  \Bigg/  
  \sqrt{ 
    \frac{ (n + m - 2) S_p^2 \big/  \sigma^2}{ (n+ m - 2) } 
    } \\
  & = \frac{U}{\sqrt{V/(n+m-2)}}
\end{align*}



## Distribution of two-sample t-statistic {.smaller}
### Proof

- By construction $\overline{X}- \overline{Y}$ is independent of $S_X^2$ and $S_Y^2$

- Therefore $\overline{X}- \overline{Y}$ is independent of $S_p^2$

- We conclude that $U$ and $V$ are independent

- In conclusion, we have shown that
$$
T = \frac{U}{\sqrt{V/(n+m-2)}} \,, \qquad 
U \sim N(0,1) \,, \qquad V \sim \chi_{n + m - 2}^2
$$

- By the Theorem in Slide 118 of Lecture 2, we conclude that
$$
T \sim t_{n+m-2}
$$




## The two-sample t-test {.smaller}

Suppose given two **independent** samples

- Sample $x_1, \ldots, x_n$ from $N(\mu_X,\sigma^2)$ of size $n$
- Sample $y_1, \ldots, y_m$ from $N(\mu_Y,\sigma^2)$ of size $m$


The two-sided hypothesis for **difference in means** is
$$
H_0 \colon \mu_X = \mu_Y \,, \quad \qquad 
H_1 \colon \mu_X \neq \mu_Y
$$

The one-sided alternative hypotheses are
$$
H_1 \colon \mu_X < \mu_Y \quad  \text{ or } \quad 
H_1 \colon \mu_X > \mu_Y
$$





## Procedure: 3 Steps {.smaller}

1. **Calculation**: Compute the two-sample t-statistic
$$
t = \frac{ \overline{x} - \overline{y}}{ s_p \ \sqrt{ \dfrac{1}{n} + \dfrac{1}{m} }}
$$
where sample means and pooled variance estimator are
$$
\overline{x} = \frac{1}{n} \sum_{i=1}^n x_i  \qquad 
\overline{y} = \frac{1}{m} \sum_{i=1}^m y_i  \qquad
s_p^2 = \frac{ (n-1) s_X^2 + (m - 1) s_Y^2 }{ m + n - 2} 
$$
$$
s_X^2 = \frac{\sum_{i=1}^n x_i^2 - n \overline{x}^2}{n-1}
\qquad 
s_Y^2 = \frac{\sum_{i=1}^m y_i^2 - m \overline{y}^2}{m-1}
$$



## {.smaller}

2. **Statistical Tables or R**: Find either
    * Critical value $t^*$ in [Table 1](files/Statistics_Tables.pdf)
    * p-value in R

<br>

3. **Interpretation**: Reject $H_0$ when either
$$
p < 0.05 \qquad \text{ or } \qquad t \in \,\,\text{Rejection Region} 
\qquad \qquad \qquad \qquad
(T \, \sim \, t_{n+m-1})
$$




| Alternative                      | Rejection Region  | $t^*$             | p-value         |
|----------------------------------|-------------------|-------------------|-----------------|
| $\mu_X \neq \mu_Y$               | $|t| > t^*$       | $t_{n+m-1}(0.025)$| $2P(T > |t|)$   |
| $\mu_X < \mu_Y$                  | $t < - t^*$       | $t_{n+m-1}(0.05)$ | $P(T < t)$      |
| $\mu_X > \mu_Y$                  | $t > t^*$         | $t_{n+m-1}(0.05)$ | $P(T > t)$      |
: {tbl-colwidths="[25,25,25,25]"}




## {.smaller}

```{r}
# Degrees of freedom
df <- 11

# Values for x-axis
x <- seq(-4, 4, length.out = 1000)

# Calculate PDF of t-distribution
pdf <- dt(x, df)

# Set layout for 3 plots side by side
par(mfrow = c(1, 3))

# Left-tailed test
plot(x, pdf, type = "l", col = "blue", lwd = 2, xlab = "", ylab = "", main = expression("Left-Tailed Test: " ~ mu[X] < mu[Y]))
x_fill_left <- x[x <= qt(0.05, df)]
y_fill_left <- pdf[x <= qt(0.05, df)]
polygon(c(x_fill_left, rev(x_fill_left)), c(y_fill_left, rep(0, length(y_fill_left))), col = "gray", border = NA)
text(qt(0.05, df) - 1.3, dt(qt(0.05, df), df) - 0.01, expression(-t(0.05)), pos = 3, col = "red", cex = 1.1)
legend("topright", legend = c("Rej. Region area = 0.05"), fill = "gray", cex = 1)

# Two-tailed test
plot(x, pdf, type = "l", col = "blue", lwd = 2, xlab = "", ylab = "", main = expression("Two-Tailed Test: " ~ mu[X] != mu[Y]))
x_fill_left <- x[x <= qt(0.025, df)]
y_fill_left <- pdf[x <= qt(0.025, df)]
polygon(c(x_fill_left, rev(x_fill_left)), c(y_fill_left, rep(0, length(y_fill_left))), col = "gray", border = NA)

x_fill_right <- x[x >= qt(0.975, df)]
y_fill_right <- pdf[x >= qt(0.975, df)]
polygon(c(x_fill_right, rev(x_fill_right)), c(y_fill_right, rep(0, length(y_fill_right))), col = "gray", border = NA)

text(qt(0.025, df) - 1.0, dt(qt(0.025, df), df) + 0.0, expression(-t(0.025)), pos = 3, col = "red", cex = 1.1)
text(qt(0.975, df) + 1.0, dt(qt(0.975, df), df) + 0.0, expression(t(0.025)), pos = 3, col = "red", cex = 1.1)
legend("topright", legend = c("Rej. Region area = 0.025 x 2"), fill = "gray", cex = 1)

# Right-tailed test
plot(x, pdf, type = "l", col = "blue", lwd = 2, xlab = "", ylab = "", main = expression("Right-Tailed Test: " ~ mu[X] > mu[Y]))
x_fill_right <- x[x >= qt(0.95, df)]
y_fill_right <- pdf[x >= qt(0.95, df)]
polygon(c(x_fill_right, rev(x_fill_right)), c(y_fill_right, rep(0, length(y_fill_right))), col = "gray", border = NA)
text(qt(0.95, df) + 1.3, dt(qt(0.95, df), df) - 0.01, expression(t(0.05)), pos = 3, col = "red", cex = 1.1)
legend("topright", legend = c("Rej. Region area = 0.05"), fill = "gray", cex = 1)

# Reset layout
par(mfrow = c(1, 1))
```


Reject $H_0$ if t-statistic $t$ falls in the *Rejection Region* (in gray). Here $t \sim t_{n+m-1}$


## The two-sample t-test in R {.smaller}
### General commands

1. Store the samples $x_1,\ldots,x_n$ and $y_1,\ldots,y_m$ in two R vectors
    * ``x <- c(x1, ..., xn)`` $\qquad$ ``y <- c(y1, ..., ym)``

2. Perform a two-sample t-test on ``x`` and ``y``
    
|Alternative                  |    R command                               |
|-----------------------------|--------------------------------------------|
|$\mu_X \neq \mu_Y$ | ``t.test(x, y, var.equal = T)``                         |
|$\mu_X < \mu_Y$    |``t.test(x, y, var.equal = T, alt = "less")`` |
|$\mu_X > \mu_Y$    |``t.test(x, y, var.equal = T, alt = "greater")`` |
: {tbl-colwidths="[25,75]"}
    
    
3. Read output: similar to one-sample t-test
    * The main quantity of interest is p-value




## Comments on command ``t.test(x, y)`` {.smaller}


1. ``mu = mu0`` tells R to test null hypothesis:
    $$
    H_0 \colon \mu_X - \mu_Y = \mu_0  \qquad \quad (\text{default is } \, \mu_0 = 0)
    $$ 


2. ``var.equal = T`` tells R to assume that populations have same variance
$$
\sigma_X^2 = \sigma^2_Y
$$

3. In this case R computes the t-statistic with formula discussed earlier
    $$
    t = \frac{ \overline{x} - \overline{y} }{s_p \sqrt{ \dfrac{1}{n} + \dfrac{1}{m} }}
    $$





## Comments on command ``t.test(x, y)`` {.smaller}

**Warning**: If ``var.equal = T`` is **not** specified then

- R assumes that populations have different variance
    $\sigma_X^2 \neq \sigma^2_Y$
- In this case the t-statistic 
  $$
    t = \frac{ \overline{x} - \overline{y} }{s_p \sqrt{ \dfrac{1}{n} + \dfrac{1}{m} }}
  $$
is **NOT** t-distributed

- R performs the **Welch t-test** instead of the **classic t-test**  
(more on this later) 







# Part 3: <br>Two-sample t-test<br>Example {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## {.smaller}



| Mathematicians |$x_1$|$x_2$|$x_3$|$x_4$|$x_5$|$x_6$|$x_7$|$x_8$|$x_9$|$x_{10}$|
|:-----------    |:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:------:|
|     Wages      | 36  |  40 | 46  | 54  |  57 | 58  |  59 |  60 |  62 |   63   |


| Accountants |$y_1$|$y_2$|$y_3$|$y_4$|$y_5$|$y_6$|$y_7$|$y_8$|$y_9$|$y_{10}$|$y_{11}$|$y_{12}$|$y_{13}$|
|:----------- |:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:------:|:------:|:------:|:------:|
|  Wages      | 37  | 37  | 42  |  44 |  46 |  48 |  54 |  56 |  59 | 60     |     60 |   64   |    64  |


<br>


- **Samples:** Wage data on 10 Mathematicians and 13 Accountants
    * Wages are independent and normally distributed
    * Populations have equal variance

- **Quesion:** Is there evidence of differences in average pay?


- **Answer:** Two-sample two-sided t-test for the hypothesis
$$
H_0 \colon \mu_X = \mu_Y \,,\qquad 
H_1 \colon \mu_X \neq \mu_Y
$$





## Calculations: First sample {.smaller}

- **Sample size:** $\ n =$ No. of Mathematicians $= 10$
- **Mean**:
$$
\bar{x} = \frac{\sum_{i=1}^n x_i}{n} 
        = \frac{36+40+46+ \ldots +62+63}{10}=\frac{535}{10}=53.5
$$

- **Variance:**
\begin{align*}
  s^2_X & = \frac{\sum_{i=1}^n x_i^2 - n \bar{x}^2}{n -1 } \\
  \sum_{i=1}^n x_i^2 & =  36^2+40^2+46^2+ \ldots +62^2+63^2 = 29435 \\
  s^2_X & =  \frac{29435-10(53.5)^2}{9} = 90.2778
\end{align*}




## Calculations: Second sample {.smaller}

- **Sample size:** $\ m =$ No. of Accountants $= 13$
- **Mean**:
$$
\bar{y} = \frac{37+37+42+ \dots +64+64}{13}
        = \frac{671}{13} = 51.6154
$$

- **Variance:**
\begin{align*}
  s^2_Y & = \frac{\sum_{i=1}^m y_i^2 - m \bar{y}^2}{m - 1} \\
  \sum_{i=1}^m y_i^2 & = 37^2+37^2+42^2+ \ldots +64^2+64^2 = 35783  \\
  s^2_Y & =  \frac{35783-13(51.6154)^2}{12} = 95.7547
\end{align*}




## Calculations: Pooled Variance {.smaller}

- **Pooled variance:**
\begin{align*}
s_p^2 & = \frac{(n-1) s_X^2 + (m-1) s_Y^2}{ n + m - 2}  \\
      & = \frac{(9) 90.2778 + (12) 95.7547 }{ 10 + 13 - 2} \\
      & = 93.40746
\end{align*}


- **Pooled standard deviation:**
$$
s_p = \sqrt{93.40746} = 9.6648
$$



## Calculations: t-statistic {.smaller}

1. **Calculation:** Compute the two-sample t-statistic

\begin{align*}
t & = \frac{\bar{x} - \bar{y} }{s_p \ \sqrt{\dfrac{1}{n}+\dfrac{1}{m}}} \\
  & = \frac{53.5 - 51.6154}{9.6648 \times \sqrt{\dfrac{1}{10}+\dfrac{1}{13}}} \\
  & = \frac{1.8846}{9.6648{\times}0.4206} \\
  & = 0.464 \,\, (3\ \text{d.p.})
\end{align*}




## Completing the t-test {.smaller}


2. **Referencing Tables**:
    * Degrees of freedom are
    $${\rm df} = n + m - 2 = 10 + 13 - 2 = 21$$
    * Find corresponding critical value in [Table 1](files/Statistics_Tables.pdf)
$$
t_{21}(0.025) = 2.08
$$
(Note the value $0.025$, since this is two-sided test)




## Completing the t-test {.smaller} 

3. **Interpretation:**
    * We have that 
      $$
      | t | = 0.464 < 2.08 = t_{21}(0.025) 
      $$
    * t falls in the *acceptance region*
    * Therefore the p-value satisfies $p>0.05$
    * There is no evidence ($p>0.05$) in favor of $H_1$
    * Hence we accept that $\mu_X = \mu_Y$

4. **Conclusion:** Average pay levels seem to be the same for both professions





## The two-sample t-test in R {.smaller}


This is a two-sided t-test with assumption of equal variance. The p-value is

$$
p = 2 P(t_{n-1} > |t|) \,, \qquad t  = \frac{\bar{x} - \bar{y} }{s_p \ \sqrt{1/n + 1/m}}
$$


```r
# Enter Wages data in 2 vectors using function c()
mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Two-sample t-test with null hypothesis mu_X = mu_Y
# and equal variance assumption. Store result in answer and print.

answer <- t.test(mathematicians, accountants, var.equal = TRUE)

print(answer)
```

- Code can be downloaded here [two_sample_t_test.R](codes/two_sample_t_test.R)




## {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Comments on output**: 

1. First line: R tells us that a Two-Sample t-test is performed
2. Second line: Data for t-test is ``mathematicians`` and ``accountants``




## {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Comments on output**: 

3. Third line:
    * The t-statistic computed is $t = 0.46359$
    * **Note**: This coincides with the one computed by hand!
    * There are $21$ degrees of freedom
    * The p-values is $p = 0.6477$




##  {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Comments on output**: 

4. Fourth line: The alternative hypothesis is that the difference in means is not zero
    * This translates to $H_1 \colon \mu_X \neq \mu_Y$
    * **Warning**: This is not saying to reject $H_0$ -- R is just stating $H_1$





##  {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Comments on output**: 

5. Fifth line: R computes a $95 \%$ confidence interval for $\mu_X - \mu_Y$ 
$$
(\mu_X - \mu_Y) \in [-6.569496, 10.338727]
$$
    * **Interpretation:** If you repeat the experiment (on new data) over and over, the interval $[a,b]$ will contain $\mu_X - \mu_Y$ about $95\%$ of the times





##  {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Comments on output**: 

6. Seventh line: R computes sample mean for the two populations
    * Sample mean for ``mathematicians`` is $53.5$
    * Sample mean for ``accountants`` is $51.61538$





##  {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform two-sample t-test with null hypothesis mu_X = mu_Y
# Specify that populations have same variance
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants, var.equal = TRUE)


# Print answer
print(answer)
```


**Conclusion**: The p-value is $p = 0.6477$

- Since $p > 0.05$ we do not reject $H_0$
- Hence $\mu_X$ and $\mu_Y$ appear to be similar
- Average pay levels seem to be the same for both professions




## Comment on Assumptions {.smaller}

The previous two-sample t-test was conducted under the following assumptions:

1. Wages data is normally distributed
2. The two populations have equal variance

Using R, we can plot the data to see if these are reasonable (*graphical exploration*)

**Warnings:**  Even if the assumptions hold

- we cannot expect the samples to be exactly normal (bell-shaped)
    * rather, look for approximate normality
- we cannot expect the sample variances to match
    * rather, look for a similar spread in the data





## Estimating the sample distribution in R {.smaller}

Suppose given a data sample stored in a vector ``z``

- If the sample is **large**, we can check normality by plotting the histogram of ``z``

- Example: ``z`` sample of size 1000 form $N(0,1)$ -- Its histogram resembles $N(0,1)$

```{r}
#| echo: true
z <- rnorm(1000)                # Sample 1000 times from N(0,1) 
hist(z, probability = TRUE)     # Plot histogram, with area scaled to 1
```


##  {.smaller}

**Drawback:** Small samples $\implies$ hard to check normality from histogram

- This is true even if the data is normal
- Example: ``z`` below is sample of size 9 from $N(0,1)$ -- But histogram not normal

```{r}
#| echo: true
z <- c(-0.78, -1.67, -0.38,  0.92, -0.58,  
       0.61, -1.62, -0.06, 0.52)           # Random from N(0,1)
hist(z, probability = TRUE)                # Histogram not normal
```


## {.smaller}

**Solution:** Suppose given iid sample ``z`` from a distribution $f$
 
- The command ``density(z)`` estimates the population distribution $f$  
(Estimate based on the *sampling distribution* of ``z`` and smoothing - Not easy task)

- Example: ``z`` as in previous slide. The plot of ``density(z)`` shows normal behavior

```{r}
#| echo: true
z <- c(-0.78, -1.67, -0.38,  0.92, -0.58,  
       0.61, -1.62, -0.06, 0.52)           # Random from N(0,1)
dz <- density(z)                           # Estimate the density of z
plot(dz)                                   # Plot the estimated density
```



## {.smaller}

The R object ``density(z)`` models a 1D function (the estimated distribution of ``z``)

- As such, it contains a grid of x values, with associated y values
    * x values are stored in vector ``density(z)$x``
    * y values are stored in vector ``density(z)$y``

- These values are useful to set the axis range in a plot

```r
dz <- density(z)

plot(dz,                        # Plot dz
     xlim = range(dz$x),        # Set x-axis range
     ylim = range(dz$y))        # Set y-axis range
```

Axes range set as the **min** and **max** values of components of ``dz``



## Checking the Assumptions on our Example {.smaller}

```r
# Compute the estimated distributions
d.math <- density(mathematicians)
d.acc <- density(accountants)

# Plot the estimated distributions

plot(d.math,                                    # Plot d.math
     xlim = range(c(d.math$x, d.acc$x)),        # Set x-axis range
     ylim = range(c(d.math$y, d.acc$y)),        # Set y-axis range
     main = "Estimated Distributions of Wages") # Add title to plot
lines(d.acc,                                    # Layer plot of d.acc
      lty = 2)                                  # Use different line style
         
legend("topleft",                               # Add legend at top-left
       legend = c("Mathematicians",             # Labels for legend
                  "Accountants"), 
       lty = c(1, 2))                           # Assign curves to legend
```

Axes range set as the **min** and **max** values of components of ``d.math`` and ``d.acc``





## {.smaller}

```{r}
# Enter the Wages data
mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)

# Compute the estimated distributions
d.math <- density(mathematicians)
d.acc <- density(accountants)

# Plot the estimated distributions

plot(d.math,                                    # Plot d.math
     xlim = range(c(d.math$x, d.acc$x)),        # Set x-axis range
     ylim = range(c(d.math$y, d.acc$y)),        # Set y-axis range
     main = "Estimated Distributions of Wages") # Add title to plot
lines(d.acc,                                    # Layer plot of d.acc
      lty = 2)                                  # Use different line style
         
legend("topleft",                               # Add legend at top-left
       legend = c("Mathematicians",             # Labels for legend
                  "Accountants"), 
       lty = c(1, 2))                           # Assign curves to legend          
```


1. Wages data looks approximately normally distributed (roughly bell-shaped)

2. The two populations have similar variance (spreads look similar)  

**Conclusion:** Two-sample t-test with equal variance is appropriate $\implies$ accept $H_0$






# Part 4: <br>The Welch t-test {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Samples with different variance {.smaller}

- We just examined the two-sample t-tests

- This assumes independent normal populations with equal variance

$$
\sigma_X^2 = \sigma_Y^2
$$

- **Question:** What happens if variances are different?

- **Answer:** Use the **Welch Two-sample t-test**
    * This is a generalization of the two-sample t-test to the case $\sigma_X^2 \neq \sigma_Y^2$
    * In R it is performed with ``t.test(x, y)``
    * Note that we are just omitting the option ``var.equal = TRUE``
    * Equivalently, you may specify ``var.equal = FALSE``




## The Welch two-sample t-test {.smaller}

- **Welch t-test** consists in computing the **Welch statistic**
$$
w = \frac{\overline{x} - \overline{y}}{ \sqrt{ \dfrac{s_X^2}{n} + \dfrac{s_Y^2}{m} } } 
$$

- If sample sizes $m,n > 5$, then $w$ is **approximately t-distributed**
    * Degrees of freedom are not integer, and depend on $S_X, S_Y, n, m$

- If variances are similar, the welch statistic is comparable to the t-statistic

$$
w \approx t : = \frac{ \overline{x} - \overline{y} }{s_p \sqrt{ \dfrac{1}{n} + \dfrac{1}{m} }}
$$



## Welch t-test Vs two-sample t-test {.smaller}

If variances are similar:

- Welch statistic and t-statistic are similar
- p-value from Welch t-test is **similar** to p-value from two-sample t-test
- Since p-values are similar, most times the 2 tests yield **same decision**
- **The tests can be used interchangeably**


If variances are very different:

- Welch statistic and t-statistic are different
- p-values from the two tests can differ a lot
- The two tests might give **different decision**
- **Wrong to apply two-sample t-test, as variances are different**




## The Welch two-sample t-test in R {.smaller}

```r
# Enter Wages data

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform Welch two-sample t-test with null hypothesis mu_X = mu_Y
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants)


# Print answer
print(answer)
```

- **Note**: 
    * This is almost the same code as in Slide 86
    * Only difference: we are omitting the option ``var.equal = TRUE`` in ``t.test``





## {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform Welch two-sample t-test with null hypothesis mu_X = mu_Y
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants)


# Print answer
print(answer)
```

**Comments on output**: 

1. First line: R tells us that a Welch Two-Sample t-test is performed
    * The rest of the output is similar to classic t-test
    * Main difference is that p-value and t-statistic differ from classic t-test




## {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform Welch two-sample t-test with null hypothesis mu_X = mu_Y
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants)


# Print answer
print(answer)
```

**Comments on output**: 

2. Third line: 
    * The Welch t-statistic is $w = 0.46546$ (standard t-test gave $t = 0.46359$)
    * Degrees of freedom are fractionary $\rm{df} = 19.795$ (standard t-test $\rm{df} = 21$) 
    * The Welch t-statistic is approximately t-distributed with $W \approx t_{19.795}$

3. Fifth line: The confidence interval for $\mu_X - \mu_Y$ is also different




##  {.smaller}

```{r}
# Enter Wages data in 2 vectors using function c()

mathematicians <- c(36, 40, 46, 54, 57, 58, 59, 60, 62, 63)
accountants <- c(37, 37, 42, 44, 46, 48, 54, 56, 59, 60, 60, 64, 64)


# Perform Welch two-sample t-test with null hypothesis mu_X = mu_Y
# Store result of t.test in answer

answer <- t.test(mathematicians, accountants)


# Print answer
print(answer)
```

**Conclusion**: The p-values obtained with the 2 tests are almost the same

- Welch t-test: p-value $= 0.6467 \qquad$ Classic t-test: p-value $= 0.6477$
- Both test: $p > 0.05$, and therefore do not reject $H_0$

- **Note:** This was expected
    * The spread of the two populations is similar $\implies \, \sigma_X^2 \approx \sigma_Y^2$ 
    * Hence, Welch t-statistic approximates t-statistic $\implies$ p-values are similar



## Exercise {.smaller}

- We compare the Effect of Two Treatments on Blood Pressure Change

- Both treatments are given to a group of patients

- Measurements of changes in blood pressure are taken after 4 weeks of treatment

- Note that changes represent both positive and negative shifts in blood pressure


<br>

|            |       |       |       |       |       |       |       |       |       |       |       |       |
|------------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| **Treat. A** | -1.9  | -2.5  | -2.1  | -2.4   | -2.6  | -1.9  |       |       |       |       |       |       |
| **Treat. B** | -1.1   | -0.9   | -1.4   | 0.2   | 0.3  | 0.6   | -5  | -2.4   | 1.5   | -2.3   | -2.8  | 2.1   |
: {tbl-colwidths="[20,7.5,7.5,7.5,7.5,7.5,7.5,7.5,7.5,7.5,7.5,7.5,7.5]"}



##  {.smaller}

```{r}
#| echo: true
# Enter changes in Blood pressure data

trA <- c(-1.9, -2.5, -2.1, -2.4, -2.6, -1.9)
trB <- c(-1.1, -0.9, -1.4, 0.2, 0.3, 0.6, -5,
         -2.4, -1.5, 2.3, -2.8, 2.1)

cat("Mean of Treatment A:", mean(trA), "Mean of Treatment B:", mean(trB))
```

<br>

- Sample means show both Treatments are effective in decreasing blood pressure

- However Treatment A seems slightly better

**Question:** Perform a t-test to see if Treatment A is better
$$
H_0 \colon \mu_A = \mu_B \, , \qquad 
H_1 \colon \mu_A < \mu_B
$$




## Solution: Estimated density of Treatment A {.smaller}

```r
plot(density(trA))    
```

```{r}
# Enter changes in Blood pressure data
trA <- c(-1.9, -2.5, -2.1, -2.4, -2.6, -1.9)

# Plot the estimated distribution
plot(density(trA))    
```


- Estimated density looks bell-shaped $\implies$ First population is normal
- Sample seems concentrated between $-3$ and $-1.5$



## Estimated density of Treatment B {.smaller}

```r
plot(density(trB))    
```

```{r}
# Enter changes in Blood pressure data
trB <- c(-1.1, -0.9, -1.4, 0.2, 0.3, 0.6, -5,
         -2.4, -1.5, 2.3, -2.8, 2.1)

# Plot the estimated distribution
plot(density(trB))
       
```


- Estimated density looks bell-shaped $\implies$ Second population is normal
- Sample seems concentrated between $-7$ and $4$ 


## Findings {.smaller}

- Both populations are normal $\implies$ t-test is appropriate
- First sample seems concentrated between $-3$ and $-1.5$
- Second sample seems concentrated between $-7$ and $4$ 
- Treatment B has larger spread
- Therefore we suspect that populations have different variance

$$
\sigma_A^2 \neq \sigma_B^2
$$

**Conclusion:**

- The Welch t-test is appropriate
- Two sample t-test would not be appropriate (as it assumes equal variance)



## Apply the Welch t-test {.smaller}

We are testing the one-sided hypothesis 

$$
H_0 \colon \mu_A = \mu_B \, , \qquad 
H_1 \colon \mu_A < \mu_B
$$

```r
# Perform Welch t-test and retrieve p-value

ans <- t.test(trA, trB, alt = "less", var.equal = F) 
ans$p.value
```

```{r}
# Enter changes in Blood pressure data
trA <- c(-1.9, -2.5, -2.1, -2.4, -2.6, -1.9)
trB <- c(-1.1, -0.9, -1.4, 0.2, 0.3, 0.6, -5,
         -2.4, -1.5, 2.3, -2.8, 2.1)

# Perform Welch t-test and retrieve p-value

ans <- t.test(trA, trB, alt = "less", var.equal = F) 
ans$p.value
```

<br>

- The p-value is $p < 0.05$


- We reject $H_0 \implies$ Treatment A is more effective 





## Two-sample t-test gives different decision {.smaller}

We are testing the one-sided hypothesis 

$$
H_0 \colon \mu_A = \mu_B \, , \qquad 
H_1 \colon \mu_A < \mu_B
$$

```r
# Perform two-sample t-test and retrieve p-value

ans <- t.test(trA, trB, alt = "less", var.equal = T) 
ans$p.value
```

```{r}
# Enter changes in Blood pressure data
trA <- c(-1.9, -2.5, -2.1, -2.4, -2.6, -1.9)
trB <- c(-1.1, -0.9, -1.4, 0.2, 0.3, 0.6, -5,
         -2.4, -1.5, 2.3, -2.8, 2.1)

# Perform two-sample t-test and retrieve p-value

ans <- t.test(trA, trB, alt = "less", var.equal = T) 
ans$p.value
```

<br>

- The p-value is $p > 0.05$

- $H_0$ cannot be rejected $\implies$ There is no evidence that Treatment A is better

- **Wrong conclusion, because two-sample t-test does not apply**



## Disclaimer {.smaller}

- The previous data was synthetic, and the background story was made up!

- Nonetheless, the example is still valid

- To construct the data, I sampled as follows
    * Treatment A: Sample of size 6 from $N(-2,1)$
    * Treatment B: Sample of size 12 from $N(-1.5,9)$

- We see that 
$$
\mu_A < \mu_B \,, \qquad 
\sigma_A^2 \neq \sigma_B^2
$$

- This tells us that:
    * We can expect that some samples will support that $\mu_A < \mu_B$
    * Two-sample t-test is inappropriate because $\sigma_A^2 \neq \sigma_B^2$


## Generating the Data {.smaller} 


```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Click here to see the code I used"

# Set seed for random generation
# This way you always get the same random numbers when
# you run this code
set.seed(21) 

repeat {
  # Generate random samples
  x <- rnorm(6, mean = -2, sd = 1)
  y <- rnorm(12, mean = -1.5, sd = 3)

  # Round x and y to 1 decimal point
  x <- round(x, 1)
  y <- round(y, 1)
  
  # Perform one-sided t-tests for alternative hypothesis mu_x < mu_y 
  ans_welch <- t.test(x, y, alt = "less", var.equal = F)
  ans_t_test <- t.test(x, y, alt = "less", var.equal = T)
   
  # Check that Welch test succeeds and two-sample test fails
  if (ans_welch$p.value < 0.05 && ans_t_test$p.value > 0.05) {
    cat("Data successfully generated!!!\n\n")
    cat("Synthetic Data TrA:", x, "\n")
    cat("Synthetic Data TrB:", y, "\n\n")
    cat("Welch t-test p-value:", ans_welch$p.value, "\n")
    cat("Two-sample t-test p-value:", ans_t_test$p.value)
    break
    }
}
```


**Method:**

- Sample the data as in previous slide (round to 1 d.p. for cleaner looking data)


- **Repeat until Welch test succeeds, and two-sample t-test fails**
$$
\text{p-value of Welch test } \, < 0.05 < \, 
\text{p-value of Two-sample t-test} 
$$




# Part 5: <br> The t-test for <br> paired samples{background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Paired samples {.smaller}

- Assume to have two sample with same size

- Sometimes the two samples depend on each other in some way


1. **Twin studies:**
    * Twins are used as pairs (to control genetic or environmental factors)
    * Example: test effectiveness of a medical treatment against placebo
    * The two samples are clearly dependent  
    (think of twins as the same person)  
    * As such, the usual two-sample t-test is not applicable  
    (because it assumes independence)



## Paired samples {.smaller}

- Assume to have two sample with same size

- Sometimes the two samples depend on each other in some way



2. **Pre-test and Post-test**
    * Measure the outcome of a certain action
    * Example: does this module work for teaching R?
    * We can assess the effectiveness of something with a pre-test and a post-test
    * The two samples are clearly dependent  
    (each individual takes a test twice)
    * As such, the usual two-sample t-test is not applicable  
    (because it assumes independence)



## The paired t-test {.smaller}

Suppose given two samples

- Sample $x_1, \ldots, x_n$ from $N(\mu_X,\sigma^2_X)$

- Sample $y_1, \ldots, y_n$ from $N(\mu_Y,\sigma^2_Y)$


The hypotheses for **difference in means** are
$$
H_0 \colon \mu_X = \mu_Y \,, \quad \qquad 
H_1 \colon \mu_X \neq \mu_Y \,, \quad \mu_X < \mu_Y \,, \quad \text{ or } \quad \mu_X > \mu_Y
$$




## The paired t-test {.smaller}


**Assumption:** The data is **paired**, meaning that the **differences**

$$
d_i = x_i - y_i \,\,  \text{ are iid} \,\, N(\mu,\sigma^2) \quad \text{where} \quad \mu := \mu_X - \mu_Y
$$

The hypotheses for the difference in means are **equivalent** to

$$
H_0 \colon \mu = 0 \,, \quad \qquad 
H_1 \colon \mu \neq 0 \,, \quad \mu < 0 \,, \quad \text{ or } \quad \mu > 0
$$

**These can be tested with a one-sample t-test**


<br>

**R commands:** The paired t-test can be called with the equivalent commands

- ``t.test(x, y, paired = TRUE)`` $\qquad \quad H_0 \colon  \mu_X = \mu_Y$

- ``t.test(x - y)`` $\qquad \qquad \qquad \qquad\qquad H_0 \colon \mu = 0$




## Example 1: The 2008 crisis (again!) {.smaller}


| Month                    | J | F | M | A | M | J | J | A | S | O | N | D |
|:------------------------:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
| CCI 2007                 |86 | 86| 88| 90| 99| 97| 97| 96| 99| 97| 90| 90|
| CCI 2009                 |24 | 22| 21| 21| 19| 18| 17| 18| 21| 23| 22| 21|


- **Data:** Monthly Consumer Confidence Index (CCI) in 2007 and 2009

- **Question:** Did the crash of 2008 have lasting impact upon CCI?

- **Observations**: 
    * Data shows a massive drop in CCI between 2007 and 2009 
    * Data is clearly paired *(Pre-test and Post-test situation)*

- **Method:** Use paired $t$-test to investigate drop in mean CCI

$$ 
H_0 \colon \mu_{2007} = \mu_{2009} \,, \quad H_1 \colon \mu_{2007} > \mu_{2009} 
$$




## Perform paired t-test {.smaller}

<br>

```{r}
#| echo: true

# Enter CCI data
score_2007 <- c(86, 86, 88, 90, 99, 97, 97, 96, 99, 97, 90, 90)
score_2009 <- c(24, 22, 21, 21, 19, 18, 17, 18, 21, 23, 22, 21)

# Perform paired t-test and print p-value
ans <- t.test(score_2007, score_2009, paired = T, alt = "greater")
ans$p.value
```
 
<br>

The p-value is significant: $\,\, p < 0.05$ $\, \implies \,$ reject $H_0$ $\, \implies \,$ Drop in CCI



## Warning {.smaller}

It would be **wrong** to use a two-sample t-test

- This is because the samples are paired, and hence **dependent**

- This is further supported by computing the correlation

- High correlation implies **dependence**

<br>

```r
cor(score_2007, score_2009)                # Correlation is high
```

```{r}
score_2007 <- c(86, 86, 88, 90, 99, 97, 97, 96, 99, 97, 90, 90)
score_2009 <- c(24, 22, 21, 21, 19, 18, 17, 18, 21, 23, 22, 21)
cor(score_2007, score_2009)
```




## Example 2: Water quality samples {.smaller}

- Researchers wish to measure water quality

- There are two possible tests, one less expensive than the other

- 10 water samples were taken, and each was measured both ways

|          |      |      |      |      |      |      |      |      |      |      |
|----------|------|------|------|------|------|------|------|------|------|------|
| method1  | 45.9 | 57.6 | 54.9 | 38.7 | 35.7 | 39.2 | 45.9 | 43.2 | 45.4 | 54.8 |
| method2  | 48.2 | 64.2 | 56.8 | 47.2 | 43.7 | 45.7 | 53.0 | 52.0 | 45.1 | 57.5 |


**Question:** Do the tests give the same results?

- **Observation**: The data is paired *(twin study situation)*

- **Method:** Use paired $t$-test to investigate equality of results

$$ 
H_0 \colon \mu_1 = \mu_2  \,, \quad H_1 \colon \mu_1 \neq \mu_2  
$$



## Perform paired t-test {.smaller}

<br>

```{r}
#| echo: true

# Enter tests data
method1 <- c(45.9, 57.6, 54.9, 38.7, 35.7, 39.2, 45.9, 43.2, 45.4, 54.8)
method2 <- c(48.2, 64.2, 56.8, 47.2, 43.7, 45.7, 53.0, 52.0, 45.1, 57.5)

# Perform paired t-test and print p-value
ans <- t.test(method1, method2, paired = T)
ans$p.value
```
 

<br>

p-value is significant: $\,\, p < 0.05$ $\implies$ reject $H_0$ $\implies$ Methods perform differently




## Warning {.smaller}

It would be **wrong** to use a two-sample t-test

- This is because the samples are paired, and hence **dependent**

- This is also supported by high samples correlation

```r
cor(method1, method2)                # Correlation is very high
```

```{r}
method1 <- c(45.9, 57.6, 54.9, 38.7, 35.7, 39.2, 45.9, 43.2, 45.4, 54.8)
method2 <- c(48.2, 64.2, 56.8, 47.2, 43.7, 45.7, 53.0, 52.0, 45.1, 57.5)
cor(method1, method2) 
```



## Warning {.smaller}

In this Example, performing a two-sample t-test would lead to **wrong decision**

<br>

```r
# Perform Welch t-test and print p-value
ans <- t.test(method1, method2, paired = F)       # paired = F is default
ans$p.val
```

```{r}
# Enter tests data
method1 <- c(45.9, 57.6, 54.9, 38.7, 35.7, 39.2, 45.9, 43.2, 45.4, 54.8)
method2 <- c(48.2, 64.2, 56.8, 47.2, 43.7, 45.7, 53.0, 52.0, 45.1, 57.5)

# Perform Welch t-test and print p-value
ans <- t.test(method1, method2, paired = F)       # paired = F is default
ans$p.val
```

<br>

**Wrong conclusion:** $\,\, p > 0.05$ $\implies$ can't reject $H_0$ $\implies$ Methods perform similarly



**Bottom line:** The data is paired, therefore a paired t-test **must** be used






# Part 6: <br>More on Vectors {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## More on vectors {.smaller}

- We have seen vectors of numbers
- Further type of vectors are:
  * Character vectors
  * Logical vectors


## Character vectors {.smaller}

- A character vector is a vector of text strings
- Elements are specified and printed in quotes

```{r}
#| echo: true
x <- c("Red", "Green", "Blue")
print(x)
```

- You can use single- or double-quote symbols to specify strings
- This is as long as the left quote is the same as the right quote

```{r}
#| echo: true
x <- c('Red', 'Green', 'Blue')
print(x)
```



## Character vectors {.smaller}

``Print`` and ``cat`` produce different output on character vectors:

- ``print(x)`` prints all the strings in ``x`` separately
- ``cat(x)`` concatenates strings. There is no way to tell how many were there


::: {.column width="45%"}

```r
x <- c("Red", "Green", "Blue")
print(x)
cat(x)
```

```{r}
x <- c("Red", "Green", "Blue")
print(x)
cat(x)
```

:::



::: {.column width="45%"}

```r
y <- c("Red Green", "Blue")
print(y)
cat(y)
```

```{r}
y <- c("Red Green", "Blue")
print(y)
cat(y)
```

:::


## Logical vectors {.smaller}

- Logical vectors can take the values ``TRUE``, ``FALSE`` or ``NA``
- ``TRUE`` and ``FALSE`` can be abbreviated with ``T`` and ``F``
- ``NA`` stands for **not available**

```{r}
#| echo: true

# Create logical vector
x <- c(T, T, F, T, NA)

# Print the logical vector
print(x)
```


## Logical vectors {.smaller}

- Logical vectors are extremely useful to evaluate conditions

- Example:
  * given a numerical vector ``x``
  * we want to count how many entries are above a value ``t``


```r
# Generate a vector containing sequence 1 to 8
x <- seq(from = 1 , to = 8, by = 1)

# Generate vector of flags for entries strictly above 5
y <- ( x > 5 )

cat("Vector x is: (", x, ")")
cat("Entries above 5 are: (", y, ")")
```


```{r}
# Generate a vector containing sequence 1 to 8
x <- seq(from = 1 , to = 8, by = 1)

# Generate vector of flags for entries strictly above 5
y <- ( x > 5 )

cat("Vector x is: (", x, ")")
cat("Entries above 5 are: (", y, ")")
```



## Logical vectors -- Application {.smaller}

- Generate a vector of 1000 numbers from $N(0,1)$
- Count how many entries are above the mean $0$
- Since there are many (1000) entries, we expect a result close to 500
    * This is because *sample mean* converges to *true mean* $0$

**Question**: How to do this?

**Hint**: ``T/F`` are interpreted as ``1/0`` in arithmetic operations


::: {.column width="45%"}
```{r}
#| echo: true
T + T
T + F
```

:::

::: {.column width="45%"}
```{r}
#| echo: true
F + F
F + T + 3
```

:::



## Logical vectors -- Application {.smaller}

- The function ``sum(x)`` sums the entries of a vector ``x``
- We can use ``sum(x)`` to count the number of ``T`` entries in a logical vector ``x``



```r
x <- rnorm(1000)       # Generates vector with 1000 normal entries

y <- (x > 0)           # Generates logical vector of entries above 0

above_zero <- sum(y)   # Counts entries above zero

cat("Number of entries which are above the average 0 is", above_zero)
cat("This is pretty close to 500!")
```


```{r}
# Generate vector with 1000 normal entries
x <- rnorm(1000)

# Generate logical vector of entries above 0
y <- (x > 0)

# Count entries above zero
above_zero <- sum(y)

cat("Number of entries which are above the average 0 is", above_zero)
cat("This is pretty close to 500!")
```



## Missing values {.smaller}

- In practical data analysis, a data point is frequently unavailable
- Statistical software needs ways to deal with this
- R allows vectors to contain a special ``NA`` value - Not Available
- ``NA`` is carried through in computations: operations on ``NA`` yield ``NA`` as the result


```{r}
#| echo: true

2 * NA
NA + NA
T + NA
```




## Indexing vectors {.smaller}

- Components of a vector can be retrieved by **indexing**

- ``vector[k]`` returns k-th component of ``vector``

<br>

```{r}
#| echo: true

vector <- c("Cat", "Dog", "Mouse")

second_element <- vector[2]           # Access 2nd entry of vector

print(second_element)
```




## Replacing vector elements {.smaller}

To modify an element of a vector use the following:

- ``vector[k] <- value`` stores ``value`` in k-th component of ``vector``

<br>

```{r}
#| echo: true

vector <- c("Cat", "Dog", "Mouse")

# We replace 2nd entry of vector with string "Horse"
vector[2] <- "Horse"

print(vector)
```




## Vector slicing {.smaller}

Returning multiple items of a vactor is known as **slicing**

- ``vector[c(k1, ..., kn)]`` returns components ``k1, ..., kn``
- ``vector[k1:k2]`` returns components ``k1`` to ``k2``

```{r}
#| echo: true

vector <- c(11, 22, 33, 44, 55, 66, 77, 88, 99, 100)

# We store 1st, 3rd, 5th entries of vector in slice
slice <- vector[c(1, 3, 5)]   

print(slice)

```


## Vector slicing {.smaller}

```{r}
#| echo: true

vector <- c(11, 22, 33, 44, 55, 66, 77, 88, 99, 100)

# We store 2nd to 7th entries of vector in slice
slice <- vector[2:7]

print(slice)
```



## Deleting vector elements {.smaller}

- Elements of a vector ``x`` can be deleted by using
    * ``x[ -c(k1, ..., kn) ]`` which deletes entries ``k1, ..., kn``
    
```r
# Create a vector x
x <- c(11, 22, 33, 44, 55, 66, 77, 88, 99, 100)

# Print vector x
cat("Vector x is:", x)

# Delete 2nd, 3rd and 7th entries of x
x <- x[ -c(2, 3, 7) ]

# Print x again
cat("Vector x with 2nd, 3rd and 7th entries removed:", x)
```


```{r}
# Create a vector x
x <- c(11, 22, 33, 44, 55, 66, 77, 88, 99, 100)

# Print vector x
cat("Vector x is:", x)

# Delete 2nd, 3rd and 7th entries of x
x <- x[ -c(2, 3, 7) ]

# Print x again
cat("Vector x with 2nd, 3rd and 7th entries removed:", x)
```



## Logical Subsetting {.smaller}


- You can index or slice vectors by entering explicit indices
- You can also index vectors, or **subset**, by using **logical flag vectors**:
    * Element is extracted if corresponding entry in the flag vector is TRUE 
    * Logical flag vectors should be the same length as vector to subset

**Code**: Suppose given a vector ``x``

- Create a flag vector by using 
    * ``flag <- condition(x)``

- ``condition()`` is any function which returns ``T/F`` vector of same length as ``x``

- Subset ``x`` by using 
    * ``x[flag]``  




## Logical Subsetting {.smaller}
### Example

- The following code extracts negative components from a numeric vector
- This can be done by using
    * ``x[ x < 0 ]``

```r
# Create numeric vector x
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Get negative components from x and store them in neg_x
neg_x <- x[ x < 0 ]

cat("Vector x is:", x)
cat("Negative components of x are:", neg_x)
```


```{r} 
# Create numeric vector x
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Get negative components from x and store them in neg_x
neg_x <- x[ x < 0 ]

cat("Vector x is:", x)
cat("Negative components of x are:", neg_x)
```



## Logical Subsetting {.smaller}
### Example

- The following code extracts components falling between ``a`` and ``b``
- This can be done by using logical operator and ``&``
    * ``x[ (x > a) & (x < b) ]``

```r
# Create numeric vector
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Get components between 0 and 100
range_x <- x[ (x > 0) & (x < 100) ]

cat("Vector x is:", x)
cat("Components of x between 0 and 100 are:", range_x)
```


```{r} 
# Create numeric vector
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Get components between 0 and 100
range_x <- x[ (x > 0) & (x < 100) ]

cat("Vector x is:", x)
cat("Components of x between 0 and 100 are:", range_x)
```






## The function Which {.smaller}

- ``which()`` allows to convert a logical vector ``flag`` into a numeric index vector 
    * ``which(flag)`` is vector of indices of ``flag`` which correspond to ``TRUE``


```r
# Create a logical flag vector
flag <- c(T, F, F, T, F)

# Indices for  flag which
true_flag <- which(flag)

cat("Flag vector is:", flag)
cat("Positions for which Flag is TRUE are:", true_flag)
```


```{r}
# Create a logical flag vector
flag <- c(T, F, F, T, F)

# Indices for  flag which
true_flag <- which(flag)

cat("Flag vector is:", flag)
cat("Positions for which Flag is TRUE are:", true_flag)
```



## The function Which -- Application {.smaller}

``which()`` can be used to delete certain entries from a vector ``x``

- Create a flag vector by using 
    * ``flag <- condition(x)``

- ``condition()`` is any function which returns ``T/F`` vector of same length as ``x``

- Delete entries flagged by ``condition`` using the code
    * ``x[ -which(flag) ]``




## The function Which -- Application {.smaller}
### Example

```r
# Create numeric vector x
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Print x
cat("Vector x is:", x)

# Flag positive components of x
flag_pos_x <- (x > 0)

# Remove positive components from x
 x <- x[ -which(flag_pos_x) ]

# Print x again
cat("Vector x with positive components removed:", x)
```


```{r} 
# Create numeric vector x
x <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)

# Print x
cat("Vector x is:", x)

# Flag positive components of x
flag_pos_x <- (x > 0)

# Remove positive components from x
 x <- x[ -which(flag_pos_x) ]

# Print x again
cat("Vector x with positive components removed:", x)
```





## Functions that create vectors {.smaller}

The main functions to generate vectors are

- ``c()`` concatenate
- ``seq()`` sequence
- ``rep()`` replicate

We have already met ``c()`` and ``seq()`` but there are more details to discuss



## Concatenate {.smaller}

**Recall:** ``c()`` generates a vector containing the input values

```{r}
#| echo: true

# Generate a vector of values 1, 2, 3, 4, 5
x <- c(1, 2, 3, 4, 5)

# Print the vector
print(x)
```


## Concatenate {.smaller}

- ``c()`` can also concatenate vectors 
- This was you can **add** entries to an **existing vector**

```{r}
#| echo: true

# Create 2 vectors
x <- c(1, 2, 3, 4, 5)
y <- c(6, 7, 8)

# Concatenate vectors x and y, and also add element 9
z <- c(x, y, 9)

# Print the resulting vector
print(z)
```




## Concatenate {.smaller}


- You can assign names to vector elements

- This modies the way the vector is printed

```{r}
#| echo: true

# We specify a vector with 3 named entries
x <- c(first = "Red", second = "Green", third = "Blue")

# Print the named vector
print(x)
```


## Concatenate {.smaller}

Given a named vector ``x``

- Names can be extracted with ``names(x)``
- Values can be extracted with ``unname(x)``

```r
# Create named vector
x <- c(first = "Red", second = "Green", third = "Blue")

# Access names of x via names(x)
names_x <- names(x)

# Access values of x via unname(x)
values_x <- unname(x)

cat("Names of x are:", names(x))
cat("Values of x are:", unname(x))
```


```{r}
# Create named vector
x <- c(first = "Red", second = "Green", third = "Blue")

# Names of componenents of x are accessed via names(x)
names_x <- names(x)

# Values of componenents of x are accessed via unname(x)
values_x <- unname(x)

cat("Names of x are:", names(x))
cat("Values of x are:", unname(x))
```


## Concatenate {.smaller}

- All elements of a vector have the **same** type
- Concatenating vectors of different types leads to conversion


```{r}
#| echo: true

c(FALSE, 2)        # Converts FALSE to 0
```

<br>

```{r}
#| echo: true

c(pi, "stats")     # Converts pi to string 

```

<br>

```{r}
#| echo: true

c(TRUE, "stats")   # Converts TRUE to string
```



## Sequence {.smaller}

- Recall the syntax of ``seq`` is
  * ``seq(from =, to =, by =, length.out =)``

- Omitting the third argument assumes that ``by = 1``

```{r}
#| echo: true

# The following generates a vector of integers from 1 to 6
x <- seq(1, 6)

print(x)
```


## Sequence {.smaller}


- ``seq(x1, x2)`` is equivalent to ``x1:x2``
- Syntax ``x1:x2`` is **preferred** to ``seq(x1, x2)``


```r
# Generate two vectors of integers from 1 to 6
x <- seq(1, 6)
y <- 1:6

cat("Vector x is:", x)
cat("Vector y is:", y)
cat("They are the same!")
```


```{r}
# Generate two vectors of integers from 1 to 6
x <- seq(1, 6)
y <- 1:6

cat("Vector x is:", x)
cat("Vector y is:", y)
cat("They are the same!")
```






## Replicate {.smaller}

``rep`` generates repeated values from a vector:

- ``x`` vector 
- ``n`` integer
- ``rep(x, n)`` repeats ``n`` times the vector ``x``

```r
# Create a vector with 3 components
x <- c(2, 1, 3)

# Repeats 4 times the vector x
y <- rep(x, 4)

cat("Original vector is:", x)
cat("Original vector repeated 4 times:", y)
```

```{r}
# Create a vector with 3 components
x <- c(2, 1, 3)

# Repeats 4 times the vector x
y <- rep(x, 4)

cat("Original vector is:", x)
cat("Original vector repeated 4 times:", y)
```
  

## Replicate {.smaller}


The second argument of ``rep()`` can also be a vector:

- Given ``x`` and ``y`` vectors
- ``rep(x, y)`` repeats entries of ``x`` as many times as corresponding entries of ``y``


```r
x <- c(2, 1, 3)         # Vector to replicate
y <- c(1, 2, 3)         # Vector saying how to replicate 

z <- rep(x, y)          # 1st entry of x is replicated 1 time
                        # 2nd entry of x is replicated 2 times
                        # 3rd entry of x is replicated 3 times

cat("Original vector is:", x)
cat("Original vector repeated is:", z)
```


```{r}
x <- c(2, 1, 3)         # Vector to replicate
y <- c(1, 2, 3)         # Vector saying how to replicate 

z <- rep(x, y)          # 1st entry of x is replicated 1 time
                        # 2nd entry of x is replicated 2 times
                        # 3rd entry of x is replicated 3 times

cat("Original vector is:", x)
cat("Original vector repeated is:", z)
```

## Replicate {.smaller}

- ``rep()`` can be useful to create vectors of labels
- **Example**: Suppose we want to collect some numeric data on 3 Cats and 4 Dogs


```{r}
#| echo: true

x <- c("Cat", "Dog")     # Vector to replicate

y <- rep(x, c(3, 4))     # 1st entry of x is replicated 3 times
                         # 2nd entry of x is replicated 4 times

cat("Vector of labels is:", y)
```
