---
title: "Statistical Models"
subtitle: "Lecture 8"
from: markdown+emoji
author: 
    - name: Dr. Silvio Fanzon
      id: sf
      email: S.Fanzon@hull.ac.uk
      url: https://www.silviofanzon.com
      affiliations: University of Hull
---



::: {.content-hidden}
$
{{< include macros.tex >}}
$
:::




# Lecture 8: <br>The Maths of <br>Regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Outline of Lecture 8



1. Simple linear regression
2. General linear regression
3. Simple regression as general regression
4. Multiple linear regression
5. Coefficient of determination $R^2$
6. Example of multiple regression
7. Two sample t-test as general regression





# Part 1: <br>Simple linear<br>regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::








## Simple linear regression: Motivation {.smaller}

- **Model:** Suppose given two random variables $X$ and $Y$
  
  ::: {.column width="48%"}
  * $X$ models *observed values*
  :::
  ::: {.column width="48%"}
  * $Y$ models a *response*
  :::
  
- **Goal of Regression:** Learn (approximate) the conditional distribution
  $$
  Y | X
  $$
  * $Y | X$ allows to predict values of $Y$ from values of $X$
  * To determine $Y|X$, we need the **joint distribution** of $(X,Y)$

- **Problem:** The joint distribution of $(X,Y)$ is **unknown**. We have partial knowledge in the form of paired observations
  $$
  (x_1,y_1) , \ldots, (x_n,y_n) \quad \text{ observed from } \quad (X,Y)
  $$
  
- **Goal:** Use the observations (data) to learn $Y|X$





## Why not just use Least-Squares? {.smaller}


::: {.column width="50%"}

**Least-Squares:** 

- Naive solution to regression problem 

- Find a line of best fit
$$
y = \hat \alpha + \hat \beta x
$$

- Such line explains the data, i.e.,
$$
y_i \approx \hat{y}_i \,, \qquad \hat{y}_i = \hat \alpha + \hat \beta x_i
$$

:::


::: {.column width="48%"}

```{r}
#| echo: false
#| fig-asp: 1

# Read the data
x <- c(1.9, 0.8, 1.1, 0.1, -0.1, 4.4, 4.6, 1.6, 5.5, 3.4)
y <- c(0.7, -1.0, -0.2, -1.2, -0.1, 3.4, 0.0, 0.8, 3.7, 2.0)

# Create the plot
plot(x, y, xlab = "", ylab = "", pch = 16, cex = 2)

# Add labels
mtext("X", side = 1, line = 3, cex = 2.1)
mtext("Y", side = 2, line = 2.5, cex = 2.1)

# Fit linear regression
fit <- lm(y ~ x)

# Plot regression line in red
abline(fit, col = "red", lwd = 5)
```

:::



## Why not just use Least-Squares? {.smaller}

::: {.column width="50%"}

**Drawbacks of least squares:** 

- Only predicts values of $y$ such that
$$
(x,y)  \, \in \, \text{ Line}
$$


- Ignores that $(x_i,y_i)$ come from joint distribution $(X,Y)$

- We don't have meaningful estimates for the error (residuals)
$$
\e_i = y_i - \hat{y}_i \,, \qquad \hat{y}_i = \hat \alpha + \hat \beta x_i
$$


:::


::: {.column width="48%"}

```{r}
#| echo: false
#| fig-asp: 1

# Read the data
x <- c(1.9, 0.8, 1.1, 0.1, -0.1, 4.4, 4.6, 1.6, 5.5, 3.4)
y <- c(0.7, -1.0, -0.2, -1.2, -0.1, 3.4, 0.0, 0.8, 3.7, 2.0)

# Create the plot
plot(x, y, xlab = "", ylab = "", pch = 16, cex = 2)

# Add labels
mtext("X", side = 1, line = 3, cex = 2.1)
mtext("Y", side = 2, line = 2.5, cex = 2.1)

# Fit linear regression
fit <- lm(y ~ x)

# Plot regression line in red
abline(fit, col = "red", lwd = 5)
```

:::



## Linear Regression instead of Least Squares {.smaller}

::: {.column width="50%"}

**Linear Regression:** 

- Find a **regression line** 
$$
R(x) = \alpha + \beta x
$$

- $R(x)$ predicts **most likely** value of $Y$ when $X = x$

- We will see that regression line coincides with line of best fit
$$
  R(x) = \hat \alpha + \hat \beta x
$$

- **Regression gives statistical meaning to Least Squares**

:::


::: {.column width="48%"}

```{r}
#| echo: false
#| fig-asp: 1

# Read the data
x <- c(1.9, 0.8, 1.1, 0.1, -0.1, 4.4, 4.6, 1.6, 5.5, 3.4)
y <- c(0.7, -1.0, -0.2, -1.2, -0.1, 3.4, 0.0, 0.8, 3.7, 2.0)

# Create the plot
plot(x, y, xlab = "", ylab = "", pch = 16, cex = 2)

# Add labels
mtext("X", side = 1, line = 3, cex = 2.1)
mtext("Y", side = 2, line = 2.5, cex = 2.1)

# Fit linear regression
fit <- lm(y ~ x)

# Plot regression line in red
abline(fit, col = "red", lwd = 5)
```

:::




## Regression function {.smaller}
### Definition

Suppose given two random variables $X$ and $Y$

::: {.column width="30%"}

- $X$ is the **predictor**

:::

::: {.column width="49%"}

- $Y$ is the **response**

:::

::: Definition

The **regression function** of $Y$ on $X$ is the conditional expectation

$$
R \colon \R \to \R \,, \qquad \quad R(x) :=  \Expect[Y | X = x]
$$

:::



## Regression function {.smaller}
### Interpretation

::: Idea

The regression function
$$
R(x) = \Expect[Y | X = x]
$$ 
predicts the most likely value of $Y$ when we observe 
$$
X = x
$$

:::

**Notation:** We use the shorthand
$$
\Expect[Y|x] := \Expect[Y | X = x]
$$




## The regression problem {.smaller}

**Assumption:** Suppose given $n$ observations $(x_1,y_1) \,, \ldots , (x_n, y_n)$

::: {.column width="38%"}

- $x_i$ observed from $X$

:::

::: {.column width="48%"}

- $y_i$ observed from $Y$

:::


::: Problem
From $(x_1,y_1) \,, \ldots , (x_n, y_n)$, learn a regression function
$$
R(x) = \Expect[Y | x]
$$
which explains the observations, that is, such that
$$
R(x_i) = \Expect[Y | x_i] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
$$

:::

**Remember: We don't know $X$ and $Y$**




## Simple linear regression {.smaller}

- Regression problem is difficult without prior knowledge on  $\Expect[Y | x]$

- A popular model is to assume that  $\Expect[Y | x]$ is linear


::: Definition

The regression function of $Y$ on $X$ is **linear** if there exist  $\alpha$ and $\beta$ s.t.
$$
\Expect[Y | x] = \alpha +  \beta x  \,, \qquad \forall \, x \in \R
$$

:::

-  $\alpha$ and $\beta$ are called **regression coefficients**

- The above regression is called **simple** because only 2 variables are involved




## What do we mean by linear? {.smaller}

We said that the regression is **linear** if
$$
\Expect[Y | x ] = \alpha + \beta x
$$
Linearity in intended wrt the parameters $\alpha$ and $\beta$


**Examples:**

- Linear regression of $Y$ on $X^2$ is
$$
\Expect[Y | x^2 ] = \alpha + \beta x^2
$$

- Linear regression of $\log Y$ on $1/X$ is
$$
\Expect[ \log Y | x ] = \alpha + \beta \frac{1}{ x }
$$





## Simple linear regression {.smaller}
### Model Assumptions


Suppose given $n$ observations $(x_1,y_1) \,, \ldots , (x_n, y_n)$

::: {.column width="38%"}

- $x_i$ observed from $X$

:::

::: {.column width="48%"}

- $y_i$ observed from $Y$

:::



::: Definition

For each $i = 1 , \ldots, n$, we denote by $Y_i$ a random variable with distribution 
$$
Y | X = x_i
$$

:::





##  {.smaller}

**Model Assumptions:** Recall, $Y_i$ is distributed like $Y | X = x_i$

1. **Predictor is known:** The values $x_1, \ldots, x_n$ are known

2. **Normality:** The conditional distribution $Y_i$ is normal

3. **Linear mean:** There are parameters $\alpha$ and $\beta$ such that
$$
\Expect[Y_i] = \alpha + \beta x_i   \,, \qquad \forall \, i = 1, \ldots, n
$$

4. **Common variance (Homoscedasticity):** There is a parameter $\sigma^2$ such that
$$
\Var[Y_i] = \sigma^2 \,, \qquad \forall \, i = 1, \ldots, n
$$

5. **Independence:** The random variables 
$$
Y_1   \,, \ldots \,, Y_n 
$$
are independent




## Characterization of the Model {.smaller}

- Assumptions 1--5 look quite abstract
- The following Proposition gives a handy characterization


::: Proposition 

Assumptions 1-5 are satisfied if and only if
$$
Y_i = \alpha + \beta x_i + \e_i
$$
for some random variables
$$
\e_1 , \ldots, \e_n  \,\, \text{ iid } \,\, N(0,\sigma^2)
$$

:::

- The terms $\e_i$ are called **errors** or **residuals**




## Proof {.smaller}

- By Assumption 2, we have that $Y_i$ is normal

- By Assumptions 3-4, we have
$$
\Expect[Y_i] = \alpha + \beta x_i \,, \qquad \quad
\Var[Y_i] = \sigma^2
$$

- Therefore $\, Y_i  \sim  N(\alpha + \beta x_i, \sigma^2)$



- Define the random variables $\e_i$ by
$$
\e_i := Y_i   -  (\alpha + \beta x_i)  \quad \qquad \left( \text{which implies } \,\,
Y_i   =  \alpha + \beta x_i + \e_i \right)
$$

- By Assumption 5, we have that $Y_1,\ldots,Y_n$ are independent

- Therefore $\e_1,\ldots, \e_n$ are independent

- Since $Y_i  \sim  N(\alpha + \beta x_i, \sigma^2)$, we conclude that $\e_i$ are iid $N(0,\sigma^2)$

- Therefore, $Y_i   =  \alpha + \beta x_i + \e_i$ with  $\e_i$ iid $N(0,\sigma^2) \qquad \qquad \qquad \qquad \quad  \qquad  \square$




## Likelihood function {.smaller}

::: {style="font-size: 0.92em"}

Let $X_1, \ldots, X_n$ be continuous rv with joint pdf 
$$
f = f(x_1, \ldots, x_n | \theta) \,, \qquad \theta \in \Theta \,\, \, \text{ parameter}
$$ 

::: Definition

The likelihood function of $(X_1, \ldots, X_n)$ for a given sample $(x_1, \ldots, x_n)$ is
$$
L \colon \Theta \to \R \,, \qquad \quad L(\theta | x_1,\ldots, x_n ) := f(x_1, \ldots, x_n | \theta)
$$

:::


::: Theorem

Suppose Assumptions 1--5 hold. The likelihood function of $(Y_1,\ldots,Y_n)$ is 

$$
L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \Bigg(   -\frac{\sum_{i=1}^n(y_i-\alpha - \beta x_i)^2}{2\sigma^2}      \Bigg)
$$

:::


:::




## Proof {.smaller}

::: {style="font-size: 0.92em"}

- By the Assumptions, we have $Y_i \sim N( \alpha + \beta x_i , \sigma^2 )$

- Therefore, the pdf of $Y_i$ is 
$$
f_{Y_i} (y_i) =  \frac{1}{\sqrt{2\pi \sigma^2}} \exp \Bigg( -\frac{(y_i-\alpha-\beta{x_i})^2}{2\sigma^2} \Bigg)
$$


- Since $Y_1,\ldots, Y_n$ are independent, we obtain
\begin{align*}
L(\alpha,\beta, \sigma^2 | y_1, \ldots,y_n) & = f(y_1,\ldots,y_n) 
                                   = \prod_{i=1}^n f_{Y_i}(y_i)  \\
                                  & = \prod_{i=1}^n 
                                  \frac{1}{\sqrt{2\pi \sigma^2}} \exp \Bigg( -\frac{(y_i-\alpha-\beta{x_i})^2}{2\sigma^2} \Bigg) \\
                                  & = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp  \Bigg(   -\frac{\sum_{i=1}^n(y_i- \alpha - \beta x_i)^2}{2\sigma^2}     \Bigg) \qquad \qquad \square
\end{align*}  

:::



## Model Summary  {.smaller}

- Simple linear regression of $Y$ on $X$ is the function
$$
\Expect[Y | x] = \alpha + \beta x
$$

- Suppose given $n$ observations $(x_1,y_1) \,, \ldots , (x_n, y_n)$

  ::: {.column width="54%"}
  * $x_i$ observed from $X$
  :::
  ::: {.column width="46%"}
  * $y_i$ observed from $Y$
  :::


- Denote by $Y_i$ the random variable $Y | X = x_i$

- We suppose that $Y_i$ has the form
$$
Y_i = \alpha + \beta x_i + \e_i \,, \qquad \e_i \, \text{ iid } \, N(0,\sigma^2)
$$

- In particular 
$$\Expect[Y|x_i] = \Expect[Y_i] = \alpha + \beta x_i$$



## The linear regression problem {.smaller}

::: Problem
Given observations $(x_1,y_1) \,, \ldots , (x_n, y_n)$, learn a linear regression function
$$
\Expect[Y | x] = \alpha + \beta x
$$
which explains the observations, that is, such that
\begin{equation} \tag{1}
\Expect[Y_i] = \Expect[Y | x_i] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
\end{equation}

:::


::: Question

How do we enforce (1)?

:::




## {.smaller}

::: {style="font-size: 0.88em"}

**Answer:** Recall condition (1) from previous slide
\begin{equation} \tag{1}
\Expect[Y_i] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
\end{equation}


- If we want (1) to hold, we need to maximize the joint probability
\begin{equation}\tag{P}
P(Y_1 \approx y_1, \ldots, Y_n \approx y_n) = 
\int_{A} f(y) \, dy \approx |A| f(y)
\end{equation}
    * where $f$ is the joint pdf of $(Y_1,\ldots, Y_n)$
    * $A$ is (small) open set in $\R^n$ containing the observations $(y_1,\ldots,y_n)$
    * |A| is the $n$-dimensional volume of $A$



- To maximize the probability, by (P), it is enough to maximize $f(y_1,\ldots,y_n)$

- Therefore, choose parameters $\hat \alpha, \hat \beta, \hat \sigma$ which **maximize the likelihood function**
$$
\max_{\alpha,\beta,\sigma} f(y_1,\ldots,y_n | \alpha,\beta, \sigma^2) = \max_{\alpha,\beta,\sigma} \ L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n )
$$
**Maximum Likelihood Estimation:** Find parameters that maximize the Likelihood, and so (P)

:::



## Maximum Likelihood Estimation {.smaller}

::: {style="font-size: 0.94em"}

::: Theorem 

Suppose Assumptions 1--5 hold. Assume given $n$ observations
$(x_1,y_1), \ldots, (x_n,y_n)$. 
The maximization problem
$$
\max_{\alpha,\beta,\sigma}  \ L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) 
$$
admits the unique solution
$$
\hat \alpha = \overline{y} - \hat  \beta \, \overline{x} \,, \qquad 
\hat \beta = \frac{S_{xy}}{S_{xx}} \,, \qquad 
\hat{\sigma}^2  = \frac{1}{n} \sum_{i=1}^n \left(  y_i - \hat \alpha - \hat \beta x_i  \right)^2
$$

:::

- The constants $\hat \alpha, \hat \beta, \hat{\sigma}^2$ are called *Maximum Likelihood Estimators (MLE)*

- **Important:** $\hat \alpha$ and $\hat \beta$ are the Least Squares coefficients!

:::


## Proof of Theorem {.smaller}

- The $\log$ function is strictly increasing. Therefore 
$$
\max_{\alpha,\beta,\sigma} \ L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) \quad \text{ is equivalent to } \quad
\max_{\alpha,\beta,\sigma} \ \log L( \alpha,\beta, \sigma^2 | y_1, \ldots, y_n )
$$


- Recall that the likelihood is
$$
L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) =  \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left(   - \frac{\sum_{i=1}^n(y_i-\alpha - \beta x_i)^2}{2\sigma^2}      \right)
$$

- Hence the log--likelihood is
$$
\log L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) 
= - \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \sum_{i=1}^n(y_i-\alpha - \beta x_i)^2 }{2 \sigma^2}
$$



##  {.smaller}

- Suppose first that $\sigma$ is fixed. In this case the problem
$$
\max_{\alpha,\beta} \ \left\{ \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \sum_{i=1}^n(y_i-\alpha - \beta x_i)^2 }{2 \sigma^2} \right\}
$$
is equivalent to 
$$
\min_{\alpha, \beta} \ \sum_{i=1}^n(y_i-\alpha - \beta x_i)^2
$$

- This is the Least Squares problem! Hence the solution is
$$
\hat \alpha = \overline{y} - \hat  \beta \, \overline{x} \,, \qquad 
\hat \beta = \frac{S_{xy}}{S_{xx}}
$$



## {.smaller}

- Substituting $\hat \alpha$ and $\hat \beta$ we obtain
\begin{align*}
\max_{\alpha,\beta,\sigma} \ & \log L(\alpha,\beta, \sigma^2 | y_1, \ldots, y_n ) 
= \max_{\sigma} \ \log L(\hat \alpha, \hat \beta, \sigma^2 | y_1, \ldots, y_n ) \\[10pt]
& = \max_{\sigma} \ \left\{ - \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \sum_{i=1}^n(y_i-\hat\alpha - \hat\beta x_i)^2 }{2 \sigma^2}  \right\}
\end{align*}


- The above is a one-dimensional minimization problem in $\sigma$

- Taking the first order derivative wrt $\sigma$, it can be shown the unique solution is
$$
\hat{\sigma}^2  = \frac{1}{n} \sum_{i=1}^n \left(  y_i - \hat \alpha - \hat \beta x_i  \right)^2
$$

- This concludes the proof




## Least-squares vs Linear regression {.smaller}

::: {style="font-size: 0.94em"}

Linear regression and least-squares give seemingly the same answer

::: {.column width="73%"}

| **Least-squares line**     |  $y = \hat \alpha + \hat \beta x$               |
|:-------------------------- |:----------------------------------------------- | 
| **Linear regression line** | $\Expect[Y | x ] =  \hat \alpha + \hat \beta x$ |
: {tbl-colwidths="[60,40]"}

:::

::: {.column width="25%"}

:::

**Question:** Why did we define regression if it gives same answer as least-squares?

**Answer:** There is actually a big difference

- Least-squares line $y = \hat \alpha + \hat \beta x$ is just a geometric object
  * We don’t have meaningful estimates for the error (residuals)
  $$
  \e_i = y_i - \hat{y_i} \,, \qquad  \hat{y}_i = \hat \alpha + \hat \beta_n
  $$
  

- Regression line $\Expect[Y | x ] =  \hat \alpha + \hat \beta x$ is Statistical model for $Y|X$
  * Can quantify in probability how well the linear model fits the data

:::




# Part 2: <br>General linear<br>regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## Simple vs General regression {.smaller}

**Simple regression:** Response variable $Y$ depends on one predictor $X$

- The goal is to learn the distribution of 
  
  $$
  Y | X
  $$

- $Y | X$ allows to predict values of $Y$, knowing values of $X$



**General regression:** Response variable $Y$ depends on $p$ predictors $Z_1 , \ldots,  Z_p$

- The goal is to learn the distribution of 
$$
Y | Z_1, \ldots, Z_p
$$

- $Y | Z_1 , \dots, Z_p$ allows to predict values of $Y$, knowing values of $Z_1 , \dots, Z_p$




## General regression function {.smaller}

Suppose given random variables $Z_1 , \ldots, Z_p$ and $Y$

::: {.column width="45%"}

- $Z_1,\ldots, Z_p$ are the **predictors**

:::

::: {.column width="49%"}

- $Y$ is the **response**

:::


**Notation:** We denote points of $\R^p$ by
$$
z = (z_1, \ldots, z_p)
$$


::: Definition

The **regression function** of $Y$ on $Z_1,\ldots,Z_p$ is the conditional expectation

$$
R \colon \R^p \to \R \,, \qquad \quad R(z) :=  \Expect[Y | Z_1 = z_1, \ldots , Z_p = z_p]
$$

:::



## General regression function {.smaller}

::: Idea

The regression function
$$
R(z) =  \Expect[Y | Z_1 = z_1, \ldots , Z_p = z_p]
$$ 
predicts the most likely value of $Y$ when we observe 
$$
Z_1 = z_1, \,\,\ldots , \,\, Z_p = z_p
$$

:::


**Notation**: We use the shorthand
$$
\Expect[Y | z_1, \ldots, z_p] := \Expect[Y | Z_1 = z_1, \ldots , Z_p = z_p]
$$




## The general regression problem {.smaller}

**Assumption:** For each $i =1 , \ldots, n$ we assume given observations

::: {.column width="42%"}

- data point $(z_{i1}, \ldots, z_{ip}, y_i)$  

:::

::: {.column width="26%"}

- $z_{ij}$ is from $Z_j$

:::

::: {.column width="26%"}

- $y_i$ is from $Y$

:::



::: Problem
From the above data, learn a regression function
$$
R(z) = \Expect[Y | z_1, \ldots, z_p]
$$
which explains the observations, that is, such that
$$
\Expect[Y | z_{i1}, \ldots, z_{ip}] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
$$

:::




## General linear regression {.smaller}


**Key assumption:** The regression function $R(z) = \Expect[Y | z_1, \ldots, z_p]$ is **linear**


::: Definition

The regression of $Y$ on $Z_1 , \ldots, Z_p$ is **linear** if there exist $\beta_1, \ldots, \beta_p$ s.t.

$$
\Expect[Y | z_1, \ldots, z_p] =  \beta_1 z_1 + \ldots + \beta_p z_p  \,, \quad \forall \, z \in \R^p
$$

$\beta_1, \ldots, \beta_p$ are called **regression coefficients**
:::

**Note:** 

- In simple regression we have 2 coefficients $\alpha$ and $\beta$

- In multiple regression we have $p$ coefficients $\beta_1, \ldots, \beta_p$




## General linear regression {.smaller}
### Model Assumptions

For each $i =1 , \ldots, n$ we assume given

::: {.column width="42%"}

- data point $(z_{i1}, \ldots, z_{ip}, y_i)$  

:::

::: {.column width="26%"}

- $z_{ij}$ is from $Z_j$

:::

::: {.column width="26%"}

- $y_i$ is from $Y$

:::


::: Definition

For each $i = 1, \ldots, n$, we denote by $Y_i$ a random variable with distribution
$$
Y | Z_1 = z_{i1} , \ldots, Z_p = z_{ip}
$$

:::




## {.smaller}

**Model Assumptions:** Recall, $Y_i$ is distributed like $Y | Z_1 = z_{i1} , \ldots, Z_p = z_{ip}$

1. **Predictor is known:** The values $z_{i1}, \ldots, z_{ip}$ are known

2. **Normality:** The conditional distribution $Y_i$ is normal

3. **Linear mean:** There are parameters $\beta_1,\ldots,\beta_p$ such that
$$
\Expect[Y_i] = \beta_1 z_{i1} + \ldots + \beta_p z_{ip}  \,, \qquad \forall \, i = 1, \ldots, n 
$$


4. **Common variance (Homoscedasticity):** There is a parameter $\sigma^2$ such that
$$
\Var[Y_i] = \sigma^2 \,, \qquad \forall \, i = 1, \ldots, n
$$

5. **Independence:** The random variables 
$$
Y_1   \,, \ldots \,, Y_n 
$$
are independent




## Characterization of the Model {.smaller}

- Assumptions 1--5 look quite abstract
- The following Proposition gives a handy characterization

::: Proposition 

Assumptions 1-5 are satisfied if and only if
$$
Y_i =  \beta_1 z_{i1} + \ldots + \beta_p z_{ip}  + \e_{i}
$$
for some **errors** or **residuals**
$$
\e_1 , \ldots, \e_n  \,\, \text{ iid } \,\, N(0,\sigma^2)
$$

:::


**Proof:** Similar to the proof of Proposition in Slide 14. Omitted




## Likelihood function {.smaller}

Introduce the column vectors
$$
\beta := (\beta_1, \ldots, \beta_p)^T \,, \qquad 
y := (y_1,\ldots,y_n)^T
$$

::: Theorem

Suppose Assumptions 1--5 hold. The likelihood function of $(Y_1,\ldots,Y_n)$ is
$$
L(\beta, \sigma^2 | y ) = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left(   -\frac{\sum_{i=1}^n(y_i- \beta_1 z_{i1} - \ldots - \beta_p z_{ip})^2}{2\sigma^2}  \right)
$$

:::



## Proof {.smaller}

- By the Assumptions, we have $Y_i \sim N(\beta_1 z_{i1} + \ldots + \beta_p z_{ip} , \sigma^2 )$

- Therefore, the pdf of $Y_i$ is 
$$
f_{Y_i} (y_i) =  \frac{1}{\sqrt{2\pi \sigma^2}} \exp \left( -\frac{(y_i - \beta_1 z_{i1} - \ldots - \beta_p z_{ip})^2}{2\sigma^2} \right)
$$


- Since $Y_1,\ldots, Y_n$ are independent, we obtain
\begin{align*}
L(\beta, \sigma^2 | y) & = f(y_1,\ldots,y_n) 
                                    = \prod_{i=1}^n f_{Y_i}(y_i)  \\
                                  & = \prod_{i=1}^n 
                                  \frac{1}{\sqrt{2\pi \sigma^2}} \exp \left( -\frac{(y_i - \beta_1 z_{i1} - \ldots - \beta_p z_{ip})^2}{2\sigma^2} \right) \\
                                  & = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left(   -\frac{\sum_{i=1}^n(y_i- \beta_1 z_{i1} - \ldots - \beta_p z_{ip})^2}{2\sigma^2}      \right) \qquad \quad \square
\end{align*}





## Design matrix {.smaller}

For each $i = 1 , \ldots, n$ suppose given $p$ observations
$$
z_{i1}, \ldots, z_{ip}
$$

::: Definition 

The **design matrix** is the $n \times p$ matrix
$$
Z :=
\left(
\begin{array}{ccc}
z_{11} & \ldots & z_{1p} \\
z_{21} & \ldots & z_{2p} \\
\ldots & \ldots & \ldots \\
z_{n1} & \ldots & z_{np} \\
\end{array}
\right)
$$

:::







## Vectorial notation {.smaller}

\begin{align*}
y - Z\beta & = 
\left(
\begin{array}{c}
y_1 \\
y_2 \\
\ldots \\
y_n \\
\end{array}
\right)
- 
\left(
\begin{array}{ccc}
z_{11} & \ldots & z_{1p} \\
z_{21} & \ldots & z_{2p} \\
\ldots & \ldots & \ldots \\
z_{n1} & \ldots & z_{np} \\
\end{array}
\right)
\left(
\begin{array}{c}
\beta_1 \\
\ldots \\
\beta_p \\
\end{array}
\right)
\\[20pt]
& = 
\left(
\begin{array}{c}
y_1 \\
y_2 \\
\ldots \\
y_n \\
\end{array}
\right)
- 
\left(
\begin{array}{c}
\beta_1 z_{11} + \ldots + \beta_p z_{1p}\\
\beta_1 z_{21} + \ldots + \beta_p z_{2p} \\
\ldots \\
\beta_1 z_{n1} + \ldots + \beta_p z_{np} \\
\end{array}
\right) \\[20pt]
& =
\left(
\begin{array}{c}
y_1 - \beta_1 z_{11} - \ldots - \beta_p z_{1p}\\
y_2 - \beta_1 z_{21} - \ldots - \beta_p z_{2p} \\
\ldots \\
y_n - \beta_1 z_{n1} - \ldots - \beta_p z_{np} \\
\end{array}
\right) \, \in \, \R^n
\end{align*}



## RSS: Residual Sum of Squares {.smaller}

From the previous calculation, we get

$$
(y - Z \beta)^T \, (y - Z\beta) =
 \sum_{i=1}^n \left( y_i - \beta_1 z_{i1} - \ldots - \beta_p z_{ip}   \right)^2
$$


::: Definition 

Assume given

::: {.column width="42%"}

- $n \times p$ design matrix $Z$

:::

::: {.column width="26%"}

- vector $y \in \R^n$

:::


The **residual sum of squares** associated to the parameter $\beta$ is the function
$$
\RSS \colon \R^p \to \R \,, \qquad 
\RSS (\beta) :=  (y - Z \beta)^T \, (y - Z\beta)
$$

:::



## Likelihood function {.smaller}
### Vectorial notation

Using vectorial notation, the likelihood function can be re-written as

\begin{align*}
L(\beta, \sigma^2 | y ) & = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left(   -\frac{\sum_{i=1}^n(y_i - \beta_1 z_{i1} - \ldots - \beta_p z_{ip})^2}{2\sigma^2}  \right) \\[20pt]
 & = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left( - \frac{ (y - Z \beta)^T \, (y - Z\beta) }{2 \sigma^2}  \right) \\[20pt]
 & = \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left( - \frac{\RSS(\beta) }{2 \sigma^2}  \right) 
\end{align*}





## Model Summary  {.smaller}

- General linear regression of $Y$ on $Z_1, \ldots, Z_p$ is the function
$$
\Expect[Y | z_1,\ldots,z_p] = \beta_1 z_{1} + \ldots + \beta_p z_p
$$

- For each $i=1, \ldots, n$ suppose given the observations 
    * $(z_{i1}, \ldots, z_{ip}, y_i)$ observed from $(Z_1,\ldots, Z_p,Y)$


- Denote by $Y_i$ the random variable $Y | z_{i1} ,\ldots, z_{ip}$


- We suppose that $Y_i$ has the form
$$
Y_i = \beta_1 z_{i1} + \ldots + \beta_p z_{ip} + \e_i \,, \qquad \e_i \,\, \text{ iid } \,\, N(0,\sigma^2)
$$ 

- In particular,
$$
\Expect[Y | z_{i1}, \ldots, z_{ip}] = \Expect[Y_i] = \beta_1 z_{i1} + \ldots + \beta_p z_{ip}
$$




## The general linear regression problem {.smaller}

For each $i = 1 , \ldots, n$ suppose given data points
$(z_{i1}, \ldots, z_{ip}, y_i)$

::: Problem
From the above data, learn a general linear regression function
$$
\Expect[Y | z_1, \ldots, z_p] = \beta_1 z_1 +
\ldots + \beta_p z_p
$$
which explains the observations, that is, such that
\begin{equation} \tag{2}
\Expect[Y_i] = \Expect[Y | z_{i1}, \ldots, z_{ip}] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
\end{equation}

:::


::: Question

How do we enforce (2)?

:::




## {.smaller}

**Answer:** Recall condition (2) from the previous slide
\begin{equation} \tag{2}
\Expect[Y_i] \ \approx \ y_i \,, \qquad \forall \, i = 1 , \ldots, n
\end{equation}


- If we want (2) to hold, we need to maximize the joint probability
\begin{equation}\tag{P}
P(Y_1 \approx y_1, \ldots, Y_n \approx y_n) = 
\int_{A} f(y) \, dy \approx |A| f(y)
\end{equation}
    * where $f$ is the joint pdf of $(Y_1,\ldots,Y_n)$
    * $A$ is a (small) open set in $\R^n$ containing the observations $(y_1,\ldots,y_n)$
    * $|A|$ is the n-dimensional volume of $A$

- To maximize the probability, by (P), it is enough to maximize $f(y_1,\ldots,y_n)$

- Therefore, choose parameters $\hat \beta,  \hat \sigma$ which **maximize the likelihood function**
$$
\max_{\beta,\sigma} \ f(y | \beta, \sigma^2) = 
\max_{\beta,\sigma} \ L(\beta, \sigma^2 | y )
$$





## Maximum Likelihood Estimation {.smaller}

::: Theorem 

Suppose Assumptions 1--5 hold. Assume given observations
$$
(z_{i1}, \ldots, z_{ip} ,y_i) \,, \quad \forall \, i =1 ,\ldots n \,, \qquad \text{ s.t. } \,\,
\det (Z Z^T) \neq 0 
$$
The maximization problem
$$
\max_{\beta,\sigma}  \ L(\beta, \sigma^2 | y ) 
$$
admits the unique solution
$$
\hat \beta = (Z^T Z)^{-1} Z^T y  \,, \qquad 
\hat{\sigma}^2  = \frac{1}{n} \RSS(\hat \beta)
$$
:::

- The constants $\hat \beta, \hat{\sigma}^2$ are called *Maximum Likelihood Estimators (MLE)*



## Minimizing the RSS {.smaller}

The proof of previous Theorem relies on the following Lemma

::: Lemma

Assume given observations
$$
(z_{i1}, \ldots, z_{ip} ,y_i) \,, \quad \forall \, i =1 ,\ldots n \,, \qquad \text{ s.t. } \,\,
\det (Z Z^T) \neq 0 
$$
The minimization problem
$$
\min_{\beta} \ \RSS(\beta) 
$$
admits the unique solution
$$
\hat \beta = (Z^T Z)^{-1} Z^T y 
$$

:::






## Results from matrix calculus {.smaller}

- To prove the Lemma we need some auxiliary results

- In the following, $x$ denotes an $n \times 1$ column vector

$$
x = 
\left(
\begin{array}{c}
x_1 \\
x_2 \\
\vdots \\
x_n
\end{array}
\right)
$$



## Results from matrix calculus {.smaller}
### First result

- Suppose given an $n \times 1$ column vector $a$


- Define the scalar function

$$
f \colon \R^n \to \R \,, \qquad f(x) := a^T x
$$

- Then $f$ is differentiable and the gradient is constant

$$
\nabla f (x) = a^T 
$$




## Results from matrix calculus {.smaller}
### Second result 

- Suppose given an $m \times n$ matrix $A$

- Define the vectorial affine function

$$
g \colon \R^n \to \R^m \,, \qquad g(x) := A x
$$

- Then $g$ is differentiable and the Jacobian is constant

$$
J g (x) = A 
$$




## Results from matrix calculus {.smaller}
### Third result 

- Suppose given an $n \times n$ matrix $M$

- Define the quadratic form

$$
h \colon \R^n \to \R \,, \qquad
h(x) := x^T M x
$$


- Then $h$ is differentiable and the Jacobian is

$$
J h(x) = x^T (M + M^T)
$$



## Additional results on the matrix transpose {.smaller}

1. For any $m \times n$ matrix $A$
$$
(A^T)^T= A
$$

2. A square matrix $M$ of size $n \times n$ is **symmetric** iff $M^T = M$


3. Scalars can be seen as $1 \times 1$ symmetric matrices. Therefore
$$
c^T = c \,, \qquad \forall \, c \in \R
$$

5. For any $m \times n$ matrix $A$ and $n \times p$ matrix $B$
$$
(AB)^T=B^T A^T
$$

6. The matrix $A^T A$ is symmetric by point (2)
$$
(A^T A)^T = A^T (A^T)^T = A^T A
$$



## Optimality conditions {.smaller}

We also need the $n$-dimensional version of the Lemma in Slide 103 Lecture 7

::: Lemma

Suppose $f \colon \R^n \to \R$ has positive semi-definite Hessian. They are equivalent

1. The point $\hat x$ is a minimizer of $f$, that is, such that

$$
f( \hat x ) = \min_{x \in \R^n} f(x)
$$

2. The point $\hat x$ satisfies the **optimality conditions**

$$
\nabla f (\hat x) = 0
$$

:::




## Proof: Minimizing the RSS {.smaller}

- Recall the definition of Residual Sum of Squares
$$
\RSS \colon \R^{p} \to \R \,, \qquad \RSS(\beta) = (y - Z \beta)^T (y - Z \beta)
$$

- We are finally ready to minimize the $\RSS$
$$
\min_{\beta \in \R^p} \ \RSS(\beta) 
$$

- It is not too difficult to show that $\nabla^2 \RSS$ is positive-semidefinite (omitted)


- By the Lemma on Optimality Conditions in Slide 50
$$
\beta \,\, \text{ minimizes } \,\, \RSS \qquad \iff \qquad
\nabla \RSS (\beta) = 0
$$

## {.smaller}

- To compute $\nabla \RSS$, we first expand $\RSS$
\begin{align*}
\RSS(\beta) & = (y-Z \beta)^T (y-Z \beta) \\[5pt]
            & = (y^T - (Z\beta)^T ) (y-Z \beta) \\[5pt]
            & = (y^T - \beta^T Z^T ) (y-Z \beta) \\[5pt]
            & = y^T y - y^T Z \beta - \beta^T Z^T y 
                + \beta^T Z^T Z \beta
\end{align*}

- Note that
  * $\beta$ has size $p \times 1 \quad \implies \quad \beta^T$ has size $1 \times p$
  * $Z$ has size $n \times p \quad \implies \quad Z^T$ has size $p \times n$
  * $y$ has size $n \times 1 \quad \implies \quad \beta^T Z^T y$ has size
$$
\left( 1 \times p \right) \times
\left( p \times n \right) \times
\left( n \times 1 \right)  = 1 \times 1
$$



## {.smaller}

::: {style="font-size: 0.95em"}

- $\beta^T Z^T y \,$ is a scalar $\quad \implies \quad$ it is symmetric and we get
$$
\left( \beta^T Z^T y  \right)^T = \beta^T Z^T y
$$


- Using the properties of transposition we also get
$$
\left( \beta^T Z^T y  \right)^T =  y^T Z \beta
$$

- From the last 2 equations we conclude
$$
\beta^T Z^T y = y^T Z \beta
$$


- The $\RSS$ can now be written as
\begin{align*}
\RSS(\beta) & = y^T y - y^T Z \beta 
                - \textcolor{magenta}{\beta^T Z^T y} + \beta^T Z^T Z \beta \\[5pt]
            & = y^T y - y^T Z \beta - \textcolor{magenta}{y^T Z \beta}
              + \beta^T Z^T Z \beta \\[5pt]
            & = y^T y - 2 y^T Z \beta + \beta^T Z^T Z \beta
\end{align*}

:::


##  {.smaller}

- Therefore, we have
\begin{align*}
\nabla \RSS (\beta) & = \nabla ( y^T y ) 
                      - 2  \nabla (y^T Z     \beta ) + \nabla (\beta^T Z^T Z \beta ) \\[5pt]
                    & = - 2  \nabla (y^T Z     \beta ) + \nabla (\beta^T Z^T Z \beta )
\end{align*}

- We need to compute
$$
 \nabla (y^T Z     \beta )  \qquad \quad \text{ and } \qquad \quad \nabla (\beta^T Z^T Z \beta )
$$


- Note that $y^T Z$ has dimension $\, ( 1 \times n ) \times ( n \times p ) = 1 \times p$


- Since $\beta$ has dimension $p \times 1$, we can define the function 
$$
f \colon \R^p \to \R \,, \qquad 
f(\beta) := y^T Z \beta
$$


- By the First Result on Matrix Calculus in Slide 46
$$
\nabla f  = \nabla ( \textcolor{magenta}{y^T Z} \beta) = \textcolor{magenta}{y^T Z}
$$

 

## {.smaller}


- Note that $Z^T Z$ has dimension $\, ( p \times n ) \times ( n \times p ) = p \times p$

- Since $\beta$ has dimension $p \times 1$, we can define the function 

$$
h \colon \R^p \to \R \,, \qquad 
h(\beta) := \beta^T Z^T Z \beta 
$$


- By the Third Result on Matrix Calculus in Slide 48
\begin{align*}
\nabla h  & = \nabla (\beta^T \textcolor{magenta}{Z^T Z} \beta) = \beta^T (\textcolor{magenta}{Z^T Z} + (\textcolor{magenta}{Z^T Z})^T ) \\[5pt]
& = \beta^T ( Z^T Z + Z^T Z ) = 2 \beta^T (Z^T Z)
\end{align*}


- Putting together these results we obtain

\begin{align*}
\nabla \RSS (\beta) & = - 2  \nabla (y^T Z     \beta ) + \nabla (\beta^T Z^T Z \beta ) \\[5pt]
                    & = -2 y^T Z + 2 \beta^T (Z^T Z)
\end{align*}


## {.smaller}

- We can finally solve the optimality conditions:
\begin{align*}
\nabla \RSS (\beta) = 0 \qquad & \iff \qquad 
-2 y^T Z + 2 \beta^T (Z^T Z) = 0  \\[5pt]
& \iff \qquad 
 \beta^T (Z^T Z) =  y^T Z
\end{align*}


- Transposing both sides of the last equation yields
\begin{equation} \tag{3}
 (Z^T Z) \beta = Z^T y
\end{equation}


- The matrix $Z^T Z$ is invertible, since we are assuming
$$
\det (Z^T Z) \neq 0
$$

- Multiplying by $(Z^T Z)^{-1}$ both sides of (3), we get
$$
\beta = (Z^T Z)^{-1} Z^T y
$$



##  {.smaller}

- In conclusion, we have shown that

$$
\nabla \RSS(\beta) = 0 \qquad \iff \qquad 
\beta = (Z^T Z)^{-1} Z^T y
$$

- By the Lemma on Optimality Conditions in Slide 50, we infer

$$
\beta = (Z^T Z)^{-1} Z^T y  \,\,\, \text{ minimizes } \,\,\, \RSS
$$

- This concludes the proof





## Maximum Likelihood Estimation {.smaller}

We are finally in position to prove the main Theorem. We recall the statement


::: Theorem 

Suppose Assumptions 1--5 hold. Assume given observations
$$
(z_{i1}, \ldots, z_{ip} ,y_i) \,, \quad \forall \, i =1 ,\ldots n \,, \qquad \text{ s.t. } \,\,
\det (Z Z^T) \neq 0 
$$
The maximization problem
$$
\max_{\beta,\sigma}  \ L(\beta, \sigma^2 | y ) 
$$
admits the unique solution
$$
\hat \beta = (Z^T Z)^{-1} Z^T y  \,, \qquad 
\hat{\sigma}^2  = \frac{1}{n} \RSS(\hat \beta)
$$
:::





## Proof of Theorem {.smaller}

- The $\log$ function is strictly increasing. Therefore
$$
\max_{\beta,\sigma} \ L(\beta, \sigma^2 | y )
\quad 
\text{ is equivalent to } \quad 
\max_{\beta,\sigma} \ \log L( \beta, \sigma^2 | y )
$$

- Recall that the likelihood is
$$
L(\beta, \sigma^2 | y ) =  \frac{1}{(2\pi \sigma^2)^{n/2}} \, \exp   \left(   - \frac{ \RSS(\beta) }{2\sigma^2}      \right)
$$


- Hence, the log--likelihood is
$$
\log L(\beta, \sigma^2 | y ) 
= - \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \RSS(\beta) }{2 \sigma^2}
$$



##  {.smaller}

- Suppose first that $\sigma$ is fixed. In this case the problem
$$
\max_{\beta} \ \left\{ \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \RSS(\beta) }{2 \sigma^2} \right\}
$$
is equivalent to 
$$
\min_{\beta} \ \RSS(\beta)
$$

- We have already proven that the $\RSS$ is minimized at
$$
\hat \beta = (Z^T Z)^{-1} Z^T y 
$$




## {.smaller}

- Substituting $\hat \beta$, we obtain

\begin{align*}
\max_{\beta,\sigma} \ & \log L(\beta, \sigma^2 | y ) 
= \max_{\sigma} \ \log L(\hat \beta, \sigma^2 | y ) \\[5pt]
& = \max_{\sigma} \ \left\{ - \frac{n}{2} \log (2 \pi) - \frac{n}{2} \log \sigma^2 -
\frac{ \RSS(\hat \beta) }{2 \sigma^2}  \right\}
\end{align*}

- The above is a one-dimensional minimization problem in $\sigma$

- Taking the first order derivative wrt $\sigma$, it can be shown the unique solution is
$$
\hat{\sigma}^2  = \frac{1}{n} \RSS(\hat \beta)
$$

- This concludes the proof







# Part 3: <br>Simple regression as<br> general regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## General linear regression model {.smaller}


- The general linear regression model is 

$$
Y_i = \beta_1 z_{i1} +  \ldots +  \beta_p z_{ip} + \e_i \,,  \quad \e_i \sim N(0,\sigma^2) \quad \, \forall \,  i = 1 , \ldots, n
$$

- The design matrix, and data and parameters vectors are
$$
Z =
\left(
\begin{array}{ccc}
z_{11} & \ldots & z_{1p} \\
\ldots & \ldots & \ldots \\
z_{n1} & \ldots & z_{np} \\
\end{array}
\right) \in \R^{n \times p} \,, \qquad 
y =
\left(
\begin{array}{c}
y_1   \\
\ldots  \\
y_{n} \\
\end{array}
\right)\,, \qquad 
\beta =
\left(
\begin{array}{c}
\beta_1  \\
\ldots  \\
\beta_{p} \\
\end{array}
\right)
$$


- The likelihood function for general linear regression is maximized by
$$
\hat \beta = (Z^T Z)^{-1} Z^T y
$$




## Simple linear regression model {.smaller}


- The simple linear regression model is 

$$
Y_i = \alpha + \beta x_i + \e_i \,,  \,,  \quad \e_i \sim N(0,\sigma^2) \quad \, \forall \,  i = 1 , \ldots, n
$$


- The data is given by
$$
x =
\left(
\begin{array}{c}
x_1   \\
\ldots  \\
x_{n} \\
\end{array}
\right)
\qquad \quad
y =
\left(
\begin{array}{c}
y_1   \\
\ldots  \\
y_{n} \\
\end{array}
\right)
$$



- The likelihood function for simple linear regression is maximized by
$$
\hat \alpha = \overline{y} - \hat \beta \, \overline{x} \,, 
\qquad \quad \hat \beta = \frac{S_{xy}}{S_{xx}}
$$





## Simple linear regression from general {.smaller}

- The general regression model comprises the simple regression one

- To see this, consider a general linear model with $p = 2$ parameters

$$
Y_i = \beta_1 z_{i1} + \beta_2 z_{i2} + \e_i \,, \qquad i = 1 , \ldots, n
$$

- We want to compare the above to

$$
Y_i = \alpha + \beta x_i + \e_i \,, \qquad i = 1 , \ldots, n
$$


- The 2 models are equivalent iff

$$
z_{i1} = 1 \,, \qquad z_{i2} = x_i \,, \qquad \forall \, i = 1 , \ldots, n
$$



##  {.smaller}

::: {style="font-size: 0.94em"}

Therefore, the design matrix is
$$
Z =
\left(
\begin{array}{cc}
1 & x_{1}  \\
\ldots & \ldots \\
1 & x_{n} \\
\end{array}
\right) \in \R^{n \times 2}
$$



::: Theorem

With the above definition of $Z$, it holds that
$$
(Z^T Z)^{-1} Z^T y
= 
\left( 
\begin{array}{cc}
\hat \alpha \\
\hat \beta
\end{array}
\right) \,, \qquad
\hat \alpha = \overline{y} - \hat \beta \overline{x} \,, 
\qquad \hat \beta = \frac{S_{xy}}{S_{xx}}
$$


:::


In other words, the multiple linear regression estimator
$$
(Z^T Z)^{-1} Z^T y
$$
coincides with the simple linear regression estimators $\hat \alpha, \hat \beta$

:::




## Proof: Calculating $Z^T y$ {.smaller}

::: {style="font-size: 0.94em"}

- $Z$ has size $n \times 2 \qquad \implies \qquad$  $Z^T$ has size $2 \times n$
- $y$ has size $n \times 1 \qquad \implies \qquad Z^T y$ has size $\, ( 2 \times n )  \times (n \times 1) = 2 \times 1$


\begin{align*}
Z^T y
& =
\left(
\begin{array}{cc}
    1 & x_1 \\
    \ldots & \ldots\\
    1 & x_n
\end{array}
\right)^T
\left(
  \begin{array}{l}
     y_1  \\
     \ldots\\
     y_n
\end{array}\right) \\[15pt]
& = 
\left(
\begin{array}{ccc}
1 & \ldots & 1 \\
x_1 & \ldots & x_n
\end{array}
\right)
\left(
  \begin{array}{l}
     y_1  \\
     \ldots\\
     y_n
\end{array}\right) \\[15pt]
& = 
\left(\begin{array}{c}
1 \times y_1 + 1 \times y_2 + \ldots + 1 \times y_n \\
x_1 y_1 + x_2 y_2 + \ldots + x_n y_n 
\end{array} \right) \\[15pt]
& = 
\left(
  \begin{array}{c}
     n \overline{y}  \\
     \sum_{i=1}^n x_i y_i 
\end{array}
\right)
\end{align*}

:::


## Calculating $Z^T Z$ {.smaller}

::: {style="font-size: 0.94em"}

- $Z$ has size $n \times 2 \qquad \implies \qquad$  $Z^T$ has size $2 \times n$

- Therefore $Z^T Z$ has size $\, ( 2 \times n )  \times (n \times 2) = 2 \times 2$


\begin{align*}
Z^T Z & = 
\left(
\begin{array}{cc}
    1 & x_1 \\
    \ldots & \ldots\\
    1 & x_n
\end{array}
\right)^T 
\left(
\begin{array}{cc}
    1 & x_1 \\
    \ldots & \ldots\\
    1 & x_n
\end{array}
\right) \\[15pt]
& = 
\left(
\begin{array}{ccc}
1 & \ldots & 1 \\
x_1 & \ldots & x_n
\end{array}
\right)
\left(
\begin{array}{cc}
    1 & x_1 \\
    \ldots & \ldots\\
    1 & x_n
\end{array}
\right) \\[15pt]
& = 
\left(
\begin{array}{cc}
    1 \times 1 + \ldots + 1 \times 1 & 1 \times x_1 + \ldots + 1 \times x_n \\
    x_1 \times 1 + \ldots + x_n \times 1 &  x_1 \times x_1 + \ldots + x_n \times x_n
\end{array}
\right) \\[15pt]
& = 
\left(
\begin{array}{cc}
    n & n \overline{x} \\
    n \overline{x} & \sum_{i=1}^n x_i^2
\end{array}
\right)
\end{align*}

:::



## Matrix inverse: General formula {.smaller}

- Suppose to have a $2 \times n$ matrix 

$$
M 
= 
\left(
\begin{array}{cc}
a & b \\
c & d
\end{array}
\right)
$$ 

- $M$ is invertible iff 

$$
\det M = ad - bc \neq 0
$$


- If $\det M \neq 0$, the inverse of $M$ is 

$$
M^{-1}
=
\frac{1}{ ad - bc}  
\left(
\begin{array}{cc}
d & -b \\
-c & a
\end{array}\right)
$$



## Resume the proof: Calculating $(Z^T Z)^{-1}$ {.smaller}



$$
\det  \left( Z^T Z \right)  = 
\det
\left(
\begin{array}{cc}
    n & n \overline{x} \\
    n \overline{x} & \sum_{i=1}^n x_i^2
\end{array}
\right)
=
n \sum_{i=1}^n x^2_i - n^2 \overline{x}^2
= n S_{xx}
$$


- Note: $\, S_{xx} = \sum_{i=1}^n (x_i - \overline{x})^2 = 0 \quad \iff \quad x_1 = \ldots = x_n = \overline{x}$

- WLOG we can discard the trivial case $S_{xx} = 0$ because:
  * The predictors $x_1, \ldots, x_n$ are either chosen, or random
  * If they are chosen, it makes no sense to choose them all equal
  * If they are random, they will all be equal with probability $0$


- Therefore, we have
$$
\det \left( Z^T Z \right) = n S_{xx} > 0  \quad \implies \quad Z^T Z \, \text{ is invertible}
$$





## Calculating $(Z^T Z)^{-1}$ {.smaller}

- Therefore, we assume $S_{xx} > 0$

- This way, we have 

$$
\det \left( Z^T Z \right) = n S_{xx} > 0
$$

- Thus $Z^T Z$ is invertible, with inverse

\begin{align*}
(Z^T Z)^{-1} & =
\left(
\begin{array}{cc}
    n & n \overline{x} \\
    n \overline{x} & \sum_{i=1}^n x_i^2
\end{array}
\right)^{-1} \\[15pt]
& = 
\frac{1}{n S_{xx} }
\left(
\begin{array}{cc}
\sum_{i=1}^n x^2_i & -n \overline{x}\\
-n\overline{x} & n
\end{array}
\right)
\end{align*}





## Calculating $(Z^T Z)^{-1} Z^T y$ {.smaller}

- $(Z^T Z)^{-1}$ has size $2 \times 2$

- $Z^T y$ has size $2 \times 1$

- $(Z^T Z)^{-1} Z^T y$ therefore has dimensions
$$
(2 \times 2) \times (2 \times 1) = (2 \times 1)
$$

- We expect $(Z^T Z)^{-1} Z^T y$ to contain the simple regression estimators $\hat \alpha, \hat \beta$

$$
(Z^T Z)^{-1} Z^T y
= 
\left( 
\begin{array}{cc}
\hat \alpha \\
\hat \beta
\end{array}
\right)
$$


## Calculating $(Z^T Z)^{-1} Z^T y$ {.smaller}

- We have already computed $(Z^T Z)^{-1}$ and $Z^T y$

- Their product is

$$
(Z^T Z)^{-1} Z^T y = \frac{1}{n S_{xx} }
\left(
\begin{array}{cc}
\sum_{i=1}^n x^2_i & -n \overline{x}\\
-n\overline{x} & n
\end{array}
\right) \left(
  \begin{array}{c}
     n \overline{y}  \\
     \sum_{i=1}^n x_i y_i 
\end{array}
\right)
$$



## Calculating $(Z^T Z)^{-1} Z^T y$ {.smaller}

- The first entry of $(Z^T Z)^{-1} Z^T y$ is therefore

\begin{align*}
\frac{ \sum_{i=1}^n x_i^2 n \overline{y} - n \overline{x} \sum_{i=1}^n x_i y_i  }{n S_{xx}} 
& = 
\frac{ \sum_{i=1}^n x_i^2  \overline{y} - \overline{x} \sum_{i=1}^n x_i y_i  }{ S_{xx}} \\[10pt]
& = \frac{ \sum_{i=1}^n x_i^2  \overline{y} 
 \textcolor{magenta}{- \overline{y} n \overline{x}^2 +  \overline{y} n \overline{x}^2}
 - \overline{x} \sum_{i=1}^n x_i y_i  }{ S_{xx}} \\[10pt]
 & = \overline{y} \ \frac{ \sum_{i=1}^n x_i^2 - n \overline{x}^2}{ S_{xx} } - 
 \overline{x} \ \frac{ \sum_{i=1}^n x_i y_i - n \overline{x} \overline{y} }{ S_{xx} } \\[10pt]
 & =  \overline{y} \ \frac{ S_{xx} }{ S_{xx} } - 
 \overline{x} \ \frac{ S_{xy} }{ S_{xx} }  \\[10pt]
 & = \overline{y} - \hat \beta \, \overline{x} = \hat{\alpha}
\end{align*}



## Calculating $(Z^T Z)^{-1} Z^T y$ {.smaller}

- The second entry of $(Z^T Z)^{-1} Z^T y$ is

$$
\frac{ - n \overline{x} n \overline{y} + n \sum_{i=1}^n x_i y_i}{ n S_{xx} } = 
\frac{  \sum_{i=1}^n x_i y_i - n\overline{x} \overline{y} }{ S_{xx} } 
 = \frac{ S_{xy} }{S_{xx}} 
 = \hat{\beta}
$$


- We have finally proven that

$$
(Z^T Z)^{-1} Z^T y = 
\left( 
\begin{array}{cc}
\hat \alpha \\
\hat \beta
\end{array}
\right)
$$

- **Conclusion**
  * Simple linear regression is special case of general regression with $p = 2$
  * Estimator for general regression recovers estimators for simple regression




# Part 4: <br>Multiple linear<br> regression{background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Mutiple linear regression{.smaller}


Response variable $Y$ depends on $p-1$ predictors $X_2, \ldots, X_p$

- The goal is to learn the distribution of 
$$
Y | X_2, \ldots, X_p
$$

- $Y | X_2, \ldots, X_p$ allows to predict values of $Y$, knowing values of $X_2, \ldots, X_p$


**Note:** We start counting from $2$ for notational convenience





## Mutiple linear regression model {.smaller}

**Assumption:** For each $i=1, \ldots, n$ we assume given observations

::: {.column width="42%"}

- data point $(x_{i2}, \ldots, x_{ip}, y_i)$ 

:::

::: {.column width="26%"}

- $x_{ij}$ is from $X_j$

:::

::: {.column width="26%"}

- $y_i$ is from $Y$

:::


- Denote by $Y_i$ the random variable $Y | x_{i2} ,\ldots, x_{ip}$



::: Definition

The multiple linear regression model is
$$
Y_i = \beta_1 + \beta_2 x_{i2} + \ldots + \beta_p x_{ip} + \e_i
$$
where the **residuals** satisfy
$$
\e_1,\ldots, \e_n \,\, \text{ iid } \,\, N(0,\sigma^2)
$$ 

:::

 


## Mutiple vs General linear regression {.smaller}


- Multiple linear regression is special case of general linear regression

- To see this, consider a *general linear model* with $p$ parameters

$$
Y_i = \beta_1 z_{i1} + \beta_2 z_{i2} + \ldots  + \beta_p z_{ip}  + \e_i \,, \qquad i = 1 , \ldots, n
$$

- We want to compare the above to the *multiple regression model*

$$
Y_i = \beta_1 + \beta_2 x_{i2} + \ldots + \beta_p x_{ip} + \e_i
$$


- The 2 models are equivalent iff

$$
z_{i1} = 1 \,, \qquad z_{i2} = x_{i2} \,, \qquad \ldots  \qquad  z_{ip} = x_{ip} \qquad \forall \, i = 1 , \ldots, n
$$



## Mutiple linear regression estimator {.smaller}

- Hence the design matrix $Z$ and data vector $y$ are

$$
Z =
\left(
\begin{array}{cccc}
1 & x_{12} & \ldots & x_{1p} \\
1 & x_{22} & \ldots & x_{2p} \\
\ldots & \ldots & \ldots & \ldots \\
1 & x_{n2} & \ldots & x_{np} \\
\end{array}
\right) \,, \qquad 
y = \left(
\begin{array}{c}
y_1 \\
\ldots \\
y_n \\
\end{array}
\right)
$$
**The design matrix includes a first column of 1s to account for the intercept**

- $Z$ has size $n \times p$
- $y$ has size $n \times 1$


- The **MLE** for multiple regression is therefore $\hat \beta \in \R^p$
$$
\hat \beta = (Z^T Z)^{-1} Z^T y
$$








# Part 5: <br>Coefficient of <br>determination $R^2$ {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## How good is the model? {.smaller}

- Consider the general linear regression model

$$
Y_i = \beta_1 z_{i1}  + \ldots + \beta_p \, z_{ip} + \e_i
$$

- Given the observed values $y = (y_1, \ldots, y_n)$, compute the MLE by

$$
\hat \beta = (\hat \beta_1, \ldots, \hat \beta_p ) = (Z^T Z)^{-1} Z^T y
$$

::: Problem

Quantify how well the model fits the **observed values**

$$
y_1 , \ldots, y_n
$$

:::



## {.smaller}


- The **predicted values** are
$$
\hat y_i := \Expect[ Y | z_{i1}, \ldots , z_{ip} ] 
          =  \hat \beta_1 \, z_{i1} + \ldots + \hat \beta_p \, z_{ip}  \quad \implies \quad \hat y = Z \hat \beta
$$


- Recall: the **Residual sum of squares** is
$$
\RSS := \sum_{i=1}^n ( y_i - \hat y_i )^2 
$$
$\RSS$ measures deviation of predicted values from observed values

$$
\text{Model is good } \,\, \iff \,\,  \hat{y}_i \approx y_i \, \quad \forall \, i \,\, \iff \,\, \RSS \, \text{ is small}
$$

- **However:** $\RSS$ is a relative measure of precision
    * Need to rescale $\RSS$, which will lead to the coefficient of determination $R^2$
    * To this end, we need some preliminary results



## Normal equation for General Regression {.smaller}

::: Lemma

Suppose given the general linear regression model

$$
Y_i =  \beta_1 z_{i1} + \ldots + \beta_p z_{ip}  + \e_{i}
$$
for errors $\e_1 , \ldots, \e_n$ iid $N(0,\sigma^2)$. Define the **observed errors**
$$
\hat{\e}_i = y_i - \hat{y}_i 
$$
Then the **normal equation** is satisfied
$$
\hat{\e}^T Z = \mathbf{0}
$$

:::


## Proof {.smaller}

\begin{align*}
\hat{\e}^T Z & = (y - \hat y)^T Z & \hat{\e} = y - \hat{y} \\
             & = (y - Z \hat{\beta})^T Z &  \hat{y} = Z\hat{\beta} \\
             & = (y - Z (Z^T Z)^{-1} Z^T y)^T Z & \hat{\beta} = (Z^T Z)^{-1} Z^T y \\
             & = y^T (I - Z (Z^T Z)^{-1} Z^T )^T Z   & \text{factor } \,y \\
             & = y^T (I - Z (Z^T Z)^{-1} Z^T )^T (Z^T)^T  & Z = (Z^T)^T\\
             & = y^T [  Z^T (I  - Z (Z^T Z)^{-1} Z^T ) ]^T   & (AB)^T = B^TA^T \\
             & = y^T [  Z^T - \underbrace{Z^T Z (Z^T Z)^{-1}}_{I} \ Z^T ]^T \\[10pt]
             & = y^T \left[  Z^T - Z^T  \right]^T = \mathbf{0}
\end{align*}




## Total sum of squares {.smaller}


- The **total sum of squares** is 

$$
\TSS := \sum_{i=1}^n ( y_i - \overline{y} )^2
$$


- $\TSS$ measures deviation of observed values from the grand mean 


- Hence $\TSS$ measures variability of the observed data $y_1 , \ldots, y_n$
     * If data very variable, then usually $y_i$ is far from $\overline{y} \implies \TSS$ is large
     * If data not very variable, then $y_i \approx \overline{y} \implies \TSS$ is small


**Goal:** Decompose the $\TSS$




## $\RSS$ and $\ESS$ {.smaller}

- Recall: the **Residual sum of squares** is
$$
\RSS := \sum_{i=1}^n ( y_i - \hat y_i )^2 
$$
The $\RSS$ measures deviation of predicted values from observed values


- Introduce the **explained sum of squares** by
$$
\ESS := \sum_{i=1}^n ( \hat y_i - \overline{y} )^2
$$
$\ESS$ measures deviation of predicted values from the grand mean




## Decomposition of $\TSS$ {.smaller}

::: Theorem

Suppose given the general linear regression model

$$
Y_i =  \beta_1 z_{i1} + \ldots + \beta_p z_{ip}  + \e_{i}
$$
for errors $\e_1 , \ldots, \e_n$ iid $N(0,\sigma^2)$. Then

$$
\TSS = \ESS + \RSS -  2 \overline{y} \sum_{i=1}^n \hat{\e}_i
$$
where the observed errors are
$$
\hat{\e}_i = y_i - \hat{y}_i 
$$

:::




## Proof  {.smaller}

**Trick:** Add and substract $\hat{y}_i$, then expand:

\begin{align*}
\TSS & = \sum_{i=1}^n (y_i - \overline{y})^2 \\ 
& = \sum_{i=1}^n \left[ (\hat{y}_i - \overline{y})  + (y_i - \hat{y}_i) \right]^2 \\
  & = \sum_{i=1}^n (\hat{y}_i - \overline{y})^2  + (y_i - \hat{y}_i)^2 + 2 (\hat{y}_i - \overline{y})  \underbrace{(y_i - \hat{y}_i)}_{\hat{\e}_i} \\
  & = \ESS + \RSS + 2 \sum_{i=1}^n \hat{y}_i \hat{\e}_i - 2 \overline{y} \sum_{i=1}^n \hat{\e}_i 
\end{align*}




## {.smaller}

- Recall the **normal equation**

$$
\hat{\e}^T Z = \mathbf{0}
$$

- Using again that $\hat y = Z \hat{\beta}$, we get

$$
\sum_{i=1}^n \hat{y}_i \hat{\e}_i = 
\hat{\e}^T \hat{y} = 
\hat{\e}^T Z \hat{\beta} =
\mathbf{0} \hat{\beta} = 0
$$

- From the above, we conclude the thesis:
$$
\TSS  = \ESS + \RSS + 2 \underbrace{ \sum_{i=1}^n \hat{y}_i \hat{\e}_i}_{0} - 2 \overline{y} \sum_{i=1}^n \hat{\e}_i
    = \ESS + \RSS - 2 \overline{y} \sum_{i=1}^n \hat{\e}_i 
$$





## Multiple regression: $\TSS = \ESS + \RSS$  {.smaller}

::: Theorem

Suppose given the multiple regression linear model

$$
Y_i =  \beta_1 + \beta_2 x_{i2} + \ldots + \beta_p x_{ip}  + \e_{i}
$$
for errors $\e_1 , \ldots, \e_n$ iid $N(0,\sigma^2)$. Then, the sum of errors is zero
$$
\sum_{i=1}^n \hat{\e}_i = 0 \,, 
$$
where $\hat{\e}_i = y_i - \hat{y}_i$. In particular, the following decomposition holds
$$
\TSS = \ESS + \RSS
$$

:::




## Proof  {.smaller}

- Since we are dealing with multiple linear regression, $Z$ has the form

$$
Z =
\left(
\begin{array}{cccc}
1 & x_{12} & \ldots & x_{1p} \\
1 & x_{22} & \ldots & x_{2p} \\
\ldots & \ldots & \ldots & \ldots \\
1 & x_{n2} & \ldots & x_{np} \\
\end{array}
\right) 
$$


- Therefore, we have that 

$$
\hat{\e}^T Z = 
(\hat{\e}_1 , \ldots, \hat{\e}_n) 
\left(
\begin{array}{cccc}
1 & x_{12} & \ldots & x_{1p} \\
1 & x_{22} & \ldots & x_{2p} \\
\ldots & \ldots & \ldots & \ldots \\
1 & x_{n2} & \ldots & x_{np} \\
\end{array}
\right) = \left(  \sum_{i=1}^n \hat{\e}_i , \ast , \ldots, \ast  \right)
$$


## {.smaller}

- The normal equation says that 

$$
\hat{\e}^T Z = \mathbf{0}
$$

- Therefore, we conclude that 

$$
\hat{\e}^T Z  = \left(  \sum_{i=1}^n \hat{\e}_i , \ast , \ldots, \ast  \right) = (0, 0, \ldots, 0)
$$

- From the above, we infer that the **sum of the errors is zero**

$$
\sum_{i=1}^n \hat{\e}_i = 0
$$



## {.smaller}


- Multiple linear regression is a special case of general linear regression

- Therefore, we can apply the previous Theorem to get
$$
\TSS = \ESS + \RSS -  2 \overline{y} \sum_{i=1}^n \hat{\e}_i
$$

- In the previous slide we have shown that

$$
\sum_{i=1}^n \hat{\e}_i = 0
$$


- Therefore, the thesis follows 

$$
\TSS = \ESS + \RSS -  2 \overline{y} \sum_{i=1}^n \hat{\e}_i =  \ESS + \RSS
$$







## Coefficient of determination $R^2$ {.smaller}


Consider the general linear regression model
$$
Y_i = \beta_1 z_{i1}  + \ldots + \beta_p \, z_{ip} + \e_i
$$
For observed data $y = (y_1, \ldots, y_n)$, define the predictions by
$$
\hat{y} = Z\hat{\beta}  \,, \qquad \hat{\beta} = (Z^TZ)^{-1} Z^T y
$$


::: Definition

The **coefficient of determination** is 
$$
R^2 := 1 -  \frac{ \RSS }{ \TSS } = 1 - \frac{ \sum_{i=1}^n ( y_i - \hat y_i  )^2 }{ \sum_{i=1}^n ( y_i - \overline{y} )^2 }
$$

:::





## Properties of $R^2$ {.smaller}

::: Proposition

For the general linear regression model, we have that

$$
R^2 \leq 1
$$

Moreover 

$$
R^2 = 1 \qquad \iff \qquad y_i = \hat y_i \quad \,\, \forall \, i = 1 , \ldots, n
$$

:::







## Proof {.smaller}


- Since $\RSS$ and $\TSS$ are sums of squares, we have that
$\TSS,  \, \RSS \geq 0$

- Therefore, the first part of the statemet follows immediately 
$$
R^2 := 1 - \frac{\RSS}{\TSS}  \leq 1
$$

- Moreover, recalling the definition of $\RSS$, we have that
$$
\RSS = \sum_{i=1}^n (y_i - \hat{y}_i)^2 = 0 \quad \iff \quad
\hat{y}_i = y_i \quad \forall \, i
$$


- The thesis immediately follows:
$$
R^2 = 1 \,\, \iff \,\, 1 - \frac{\RSS}{\TSS} = 1  \,\, \iff \,\, \RSS = 0 \,\, \iff \,\,
\hat{y}_i = y_i \quad \forall \, i
$$




## $R^2$ might be negative {.smaller}


::: Proposition

1. For the **general** linear regression model, it might hold that
$$
R^2 < 0
$$

2. For the **multiple** regression model, it always holds that
$$
0 \leq R^2 \leq 1
$$

:::

Bottom line:

- $R^2$ can be negative for general regression (i.e. when no intercept is present)

- $0 \leq R^2 \leq 1$ for multiple regression (i.e. when intercept is present)





## Proof {.smaller}

1. For general linear regression, the following decomposition holds (Thm Slide 85)
$$
\TSS = \ESS + \RSS -  2 \overline{y} \sum_{i=1}^n \hat{\e}_i
$$


- By definition of $R^2$, we have
$$
R^2 := 1 - \frac{ \RSS }{ \TSS } = \underbrace{\frac{\ESS}{\TSS}}_{\geq 0} - \underbrace{\frac{2 \overline{y} \sum_{i=1}^n \hat{\e}_i}{\TSS}}_{\text{can have any sign}} 
$$

- Since the last term can have any sign, it might hold that $R^2 < 0$



## {.smaller}


2. For multiple linear regression, the following decomposition holds (Thm Slide 88)
$$
\TSS = \ESS + \RSS
$$


- By definition of $R^2$, we have

$$
R^2 := 1 - \frac{ \RSS }{ \TSS } = \frac{\ESS}{\TSS} \geq 0 
$$

- Therefore, we conclude that

$$
0 \leq R^2 \leq 1
$$





## Conclusion {.smaller}

We have shown that

- $R^2 \leq 1$

- $R^2 = 1 \quad \iff \quad y_i = \hat y_i \quad \,\, \forall \, i = 1 , \ldots, n$


**Conclusion:** $R^2$ measures how well the model fits the data

$$
R^2  \, \text{ close to } \, 1 \quad \implies \quad \text{  model fits data well}
$$


<br>


**Warning:** $R^2$ can be negative for general regression (no intercept)




# Part 6: <br>Example of <br>multiple regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Example: Predicting unemployment {.smaller}

::: {.column width="55%"}

**Goal:** Predict unemployment rates

- Unemp. related to Productivity Index
- Other factors might also play a role
- These factors likely depend on the year
- Use year as surrogate for other factors


:::

::: {.column width="42%"}


| Unemp. | Product. | Year |
|:------:|:--------:|:----:|
|  3.1   |  113     | 1950 |
|  1.9   |  123     | 1951 |
|  1.7   |  127     | 1952 |
|  1.6   |  138     | 1953 |
|  3.2   |  130     | 1954 |
|  2.7   |  146     | 1955 |
|  2.6   |  151     | 1956 |
|  2.9   |  152     | 1957 |
|  4.7   |  141     | 1958 |
|  3.8   |  159     | 1959 |
: {tbl-colwidths="[30,35,35]"}

:::


## Enter the data {.smaller}


- Enter the data in 3 vectors

```r 
# Unemployment data
unemployment <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)

# Productivity data
productivity <- c(113, 123, 127, 138, 130, 146, 151, 152, 141, 159)

# Year data
year <- seq(1950, 1959, by = 1)
```




## Plotting the data {.smaller}

- Want to see if either *Productivity* or *Year* predict *Unemployment*. Therefore, plot
    * Productivity Vs Unemployment
    * Year Vs Unemployment


```r
# Set up the layout with two plot panels side by side
par(mfrow = c(1, 2))

# Scatter plot: Productivity vs Unemployment
plot(productivity, unemployment, 
     xlab = "Productivity", ylab = "Unemployment",
     main = "Productivity vs Unemployment", 
     pch = 16)

# Scatter plot: Year vs Unemployment
plot(year, unemployment, 
     xlab = "Year", ylab = "Unemployment",
     main = "Year vs Unemployment", 
     pch = 16)
```


## Output {.smaller}
### There appears to be some linear relationship

```{r}
# Unemployment data
unemployment <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)

# Productivity data
productivity <- c(113, 123, 127, 138, 130, 146, 151, 152, 141, 159)

# Year data
year <- seq(1950, 1959, by = 1)

# Set up the layout with two plot panels side by side
par(mfrow = c(1, 2))

# Scatter plot: Productivity vs Unemployment
plot(productivity, unemployment, 
     xlab = "Productivity", ylab = "Unemployment",
     main = "Productivity vs Unemployment", 
     pch = 16)

# Scatter plot: Year vs Unemployment
plot(year, unemployment, 
     xlab = "Year", ylab = "Unemployment",
     main = "Year vs Unemployment", 
     pch = 16)
```




## Plotting the data in 3D {.smaller}

To better understand the data we can do a 3D scatter plot using either

- ``scatterplot3D``
  * installed via ``install.packages("scatterplot3d")``
  * produces static plots

- ``plotly``
  * installed via ``install.packages("plotly")``
  * produces interactive plots




## Plotting the data with ``scatterplot3d`` {.smaller}



```r
# Load the scatterplot3d library
library(scatterplot3d)

# Create 3D scatter plot
scatterplot3d(x = productivity,
              y = year,
              z = unemployment, 
              main = "Productivity and Year vs Unemployment", 
              xlab = "Productivity", 
              ylab = "Year", 
              zlab = "Unemployment",
              pch = 16,
              color = "black",
              angle = 50)
```



## Plotting the data with ``scatterplot3d`` {.smaller}
### The data is clearly close to a plane

```{r}
# Unemployment data
unemployment <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)

# Productivity data
productivity <- c(113, 123, 127, 138, 130, 146, 151, 152, 141, 159)

# Year data
year <- seq(1950, 1959, by = 1)

# Load the scatterplot3d library
library(scatterplot3d)

# Create 3D scatter plot
scatterplot3d(x = productivity,
              y = year,
              z = unemployment, 
              main = "Productivity and Year vs Unemployment", 
              xlab = "Productivity", 
              ylab = "Year", 
              zlab = "Unemployment", 
              pch = 16,
              color = "black",
              angle = 50)
```




## Plotting the data with ``plotly`` {.smaller}

```r
# Load the plotly library
library(plotly)

# Create a scatter plot using plot_ly
plot_ly(x = productivity, 
        y = year, 
        z = unemployment, 
        type = "scatter3d", 
        mode = "markers",
        marker = list(size = 5, color = "black")) %>%
  layout(scene = list(xaxis = list(title = "Productivity"),
                      yaxis = list(title = "Year"),
                      zaxis = list(title = "Unemployment")),
         width = 1000, 
         height = 500)
```




## Plotting the data with ``plotly`` {.smaller}
### The data is clearly close to a plane

```{r}
# Load the plotly library
library(plotly)

# Data
unemployment <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)
productivity <- c(113, 123, 127, 138, 130, 146, 151, 152, 141, 159)
year <- seq(1950, 1959, by = 1)

# Create a scatter plot using plot_ly
plot_ly(x = productivity, 
        y = year, 
        z = unemployment, 
        type = "scatter3d", 
        mode = "markers",
        marker = list(size = 5, color = "black")) %>%
  layout(scene = list(xaxis = list(title = "Productivity"),
                      yaxis = list(title = "Year"),
                      zaxis = list(title = "Unemployment")),
         width = 1000, 
         height = 500)
```



## Setting up multiple regression model {.smaller}

- We saw that the data points
$$
( \text{Productivity}, \text{Year}, \text{Unemployment}  ) \in R^3
$$
roughly lie on a **plane**

- **Goal:** Use multiple regression to compute such plane
  * Unemployment is modelled by response variable $Y$
  * Productivity is modelled by predictor $X_2$
  * Year is modelled by predictor $X_3$
  * The multiple regression model is
  $$
  \Expect[Y | x_2, x_3] = \beta_1 + \beta_2 \, x_2 + \beta_3 \, x_3
  $$



## Exercise {.smaller}

::: {.column width="54%"}

Multiple regression model is
$$
\Expect[Y | x_2, x_3] = \beta_1 + \beta_2 \, x_2 + \beta_3 \, x_3
$$

- Enter design matrix ``Z`` into R
- Enter data ``y`` into R
- Compute estimator 
$$
\hat \beta = (Z^T Z)^{-1} Z^T y
$$
    * Matrix product in R is $\,$``%*%``
    * Transpose matrix of $\,$``A``$\,$ is $\,$``t(A)``
    * Inverse of matrix $\,$``A``$\,$ is $\,$``solve(A)``

:::

::: {.column width="45%"}

::: {style="font-size: 0.85em"}

| Unemp. |  Product.| Year |
|:------:|:--------:|:----:|
| $y_i$  | $x_{i2}$ |$x_{i3}$|
|  3.1   |  113     | 1950 |
|  1.9   |  123     | 1951 |
|  1.7   |  127     | 1952 |
|  1.6   |  138     | 1953 |
|  3.2   |  130     | 1954 |
|  2.7   |  146     | 1955 |
|  2.6   |  151     | 1956 |
|  2.9   |  152     | 1957 |
|  4.7   |  141     | 1958 |
|  3.8   |  159     | 1959 |
: {tbl-colwidths="[30,35,35]"}

:::
:::





## Solution {.smaller}

::: {.column width="54%"}

The design matrix is
$$
Z=\left(\begin{array}{cccc}
    1 & 113 & 1950 \\
    1 & 123 & 1951 \\
    1 & 127 & 1952 \\
    1 & 138 & 1953 \\
    1 & 130 & 1954 \\
    1 & 146 & 1955 \\
    1 & 151 & 1956 \\
    1 & 152 & 1957 \\
    1 & 141 & 1958 \\
    1 & 159 & 1959 \\
\end{array}\right)
$$

:::

::: {.column width="45%"}

::: {style="font-size: 0.85em"}

| Unemp. |  Product.| Year |
|:------:|:--------:|:----:|
| $y_i$  | $x_{i2}$ |$x_{i3}$|
|  3.1   |  113     | 1950 |
|  1.9   |  123     | 1951 |
|  1.7   |  127     | 1952 |
|  1.6   |  138     | 1953 |
|  3.2   |  130     | 1954 |
|  2.7   |  146     | 1955 |
|  2.6   |  151     | 1956 |
|  2.9   |  152     | 1957 |
|  4.7   |  141     | 1958 |
|  3.8   |  159     | 1959 |
: {tbl-colwidths="[30,35,35]"}

:::
:::



## Solution {.smaller}

::: {.column width="54%"}

The data vector is
$$
y = 
\left(\begin{array}{c}
3.1 \\
1.9 \\
1.7 \\
1.6 \\
3.2 \\
2.7 \\
2.6 \\
2.9 \\
4.7 \\
3.8 \\
\end{array}\right)
$$

:::

::: {.column width="45%"}

::: {style="font-size: 0.85em"}

| Unemp. |  Product.| Year |
|:------:|:--------:|:----:|
| $y_i$  | $x_{i2}$ |$x_{i3}$|
|  3.1   |  113     | 1950 |
|  1.9   |  123     | 1951 |
|  1.7   |  127     | 1952 |
|  1.6   |  138     | 1953 |
|  3.2   |  130     | 1954 |
|  2.7   |  146     | 1955 |
|  2.6   |  151     | 1956 |
|  2.9   |  152     | 1957 |
|  4.7   |  141     | 1958 |
|  3.8   |  159     | 1959 |
: {tbl-colwidths="[30,35,35]"}

:::
:::








## Solution {.smaller}

::: {style="font-size: 0.93em"}

- Data vector $y$ is stored in R as usual

```r
# Store vector y
y <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)
```

- Design matrix is stored as follows


```r
# Store design matrix Z as long vector
Z <- c(1, 113, 1950,
       1, 123, 1951,
       1, 127, 1952,
       1, 138, 1953,
       1, 130, 1954,
       1, 146, 1955,
       1, 151, 1956,
       1, 152, 1957,
       1, 141, 1958,
       1, 159, 1959)

# Make Z into a matrix with 3 columns
Z <- matrix(Z, ncol = 3, byrow = TRUE)
```
:::


## Solution {.smaller}

- We need to compute the estimator

$$
\hat \beta = (Z^T Z)^{-1} Z^T y
$$


- The matrix transpose is computed with $\,$``t(A)``
- The product of two matrices $A$ and $B$ is computed with $\,$ ``A %*% B``
- We use these two operators to compute $Z^T Z$ and store it into ``m1``


```r
# Calculate m1
m1 <- t(Z) %*% Z
```


## Solution {.smaller}


- The inverse of a matrix $A$ is computed with $\,$ ``solve(A)``
- We use this function to compute $(Z^T Z)^{-1}$ and store it into ``m2``


```r
# Calculate m2
m2 <- solve(m1)
```

- We can compute $Z^T y$ with $\,$``%*%``
- This is because R automatically interprets vectors as matrices

```r
# Calculate m3
m3 <- t(Z) %*% y
```



## Solution {.smaller}

- We compute the estimator 
$$\hat{\beta}=(Z^T Z)^{-1} Z^T y
$$
and store it as $\,$ ``beta``

```r
# Calculate estimator beta
beta <- m2 %*% m3
```

- We then print the result to screen

```r
# Print estimator beta
cat("The estimator is", beta)
```


```{r}
# Store vector y
y <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)

# Store design matrix Z as long vector
Z <- c(1, 113, 1950,
       1, 123, 1951,
       1, 127, 1952,
       1, 138, 1953,
       1, 130, 1954,
       1, 146, 1955,
       1, 151, 1956,
       1, 152, 1957,
       1, 141, 1958,
       1, 159, 1959)

# Make Z into a matrix with 3 columns
Z <- matrix(Z, ncol = 3, byrow = TRUE)

# Calculate m1
m1 <- t(Z) %*% Z

# Calculate m2
m2 <- solve(m1)

# Calculate m3
m3 <- t(Z) %*% y

# Calculate estimator beta
beta <- m2 %*% m3

# Print estimator beta
cat("The estimator is", beta)
```



## Solution {.smaller}


- The computed regression model is

\begin{align*}
\Expect[Y | x_2, x_3] & = \hat\beta_1 + \hat\beta_2 \, x_2 + \hat\beta_3 \, x_3 \\[15pt]
 & = -1271.75 + (-0.1033386) \times x_2 + 0.6594171 \times x_3 
\end{align*}

- This means 

\begin{align*}
\text{Predicted Unemployment}  = 
 -1271.75 & + (-0.1033386) \times \text{Productivity}  \\[15pt]
  & + (0.6594171) \times \text{Year}
\end{align*}


- Previous code can be downloaded here [multiple_regression.R](codes/multiple_regression.R)



## Plot of data and regression plane {.smaller}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Click here to show the full code"

# Load the plotly library
library(plotly)

# Data
unemployment <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)
productivity <- c(113, 123, 127, 138, 130, 146, 151, 152, 141, 159)
year <- seq(1950, 1959, by = 1)

# Generate a grid of x and y values
x_grid <- seq(min(productivity), max(productivity), length.out = 50)
y_grid <- seq(min(year), max(year), length.out = 50)
grid <- expand.grid(productivity = x_grid, year = y_grid)

# Calculate z values for the regression plane
z_values <- -1271.75 - 0.1033386 * grid$productivity + 0.6594171 * grid$year

# Create a scatter plot using plot_ly
plot_ly(x = productivity, 
        y = year, 
        z = unemployment, 
        type = "scatter3d", 
        mode = "markers",
        marker = list(size = 5, color = "black")) %>%
  layout(scene = list(xaxis = list(title = "Productivity"),
                      yaxis = list(title = "Year"),
                      zaxis = list(title = "Unemployment")),
         width = 1000, 
         height = 500) %>%
         add_surface(
         x = x_grid, 
         y = y_grid, 
         z = matrix(z_values, nrow = length(x_grid)),
         opacity = 0.5, 
         colorscale = list(c(0, 'rgb(255,255,255)'), c(1, 'rgb(255,0,0)')))
```




## Computing the $R^2$ coefficient  {.smaller} 


- How well does the multiple linear model fit our data?

- To answer this question we compute the coefficient of determination

$$
R^2 = 1 - \frac{\RSS}{\TSS}
$$



## Computing $\TSS$  {.smaller} 

- We have that 

$$
\TSS := \sum_{i=1}^n (y_i - \overline{y})^2
$$

- We compute $\overline{y}$ with the R command

```r
# Compute mean of y
y.bar <- mean(y)
```

- We compute $\TSS$ with the R command


```r
# Compute TSS
TSS <- sum( ( y - y.bar) ^ 2 )
```


## Computing predictions  {.smaller} 

- We have that 

$$
\RSS := \sum_{i=1}^n (y_i - \hat y_i)^2
$$


- We need to compute the predictions

$$
\hat y_i :=  \hat \beta_1 + \hat \beta_2 x_{i2} + \hat \beta_3 x_{i3}
$$





## Computing predictions  {.smaller}


- $\hat y$, the design matrix $Z$, and $\hat \beta$ are

$$
\hat y = 
\left(\begin{array}{c}
    \hat y_1 \\
    \ldots \\
    \hat y_i  \\
    \ldots  \\
    \hat y_n \\
\end{array}\right) \,, \qquad 
Z = 
\left(\begin{array}{cccc}
    1 & x_{11} & x_{12} \\
    \ldots & \ldots & \ldots \\
    1 & x_{i1} & x_{i2} \\
    \ldots & \ldots & \ldots \\
    1 & x_{n1} & x_{n2} \\
\end{array}\right) \,, \qquad 
\hat \beta = 
\left(\begin{array}{c}
    \hat \beta_1 \\
    \hat \beta_2  \\
    \hat \beta_3 \\
\end{array}\right) 
$$


- Therefore, we can compute 
$$
\hat y_i =  \hat \beta_1 + \hat \beta_2 x_{i2} + \hat \beta_3 x_{i3}
$$
with the product 
$$
\hat y = Z \hat \beta
$$





## Computing $\RSS$  {.smaller}


- Recall that $\hat \beta$ is stored in ``beta``

- Compute predictions $\hat y = Z \hat{\beta}$

```r
# Compute predictions y.hat
y.hat <- Z %*% beta
```


- Recall that 
$$
\RSS = \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

- This is computed in R with

```r
# Compute RSS
RSS <- sum( (y - y.hat) ^ 2 )
```



## Computing $R^2$  {.smaller}


- The coefficient $R^2$ is computed with

```r 
# Compute coefficient of determination R^2
R2 <- 1 - RSS / TSS

# Print R^2
cat("\nThe coefficient of determination R^2 is", R2)
```

- The full code can be downloaded here [R2_multiple_regression.R](codes/R2_multiple_regression.R)





## Output & Conclusions  {.smaller}

```{r} 
# Store vector y
y <- c(3.1, 1.9, 1.7, 1.6, 3.2, 2.7, 2.6, 2.9, 4.7, 3.8)

# Store design matrix Z as long vector
Z <- c(1, 113, 1950,
       1, 123, 1951,
       1, 127, 1952,
       1, 138, 1953,
       1, 130, 1954,
       1, 146, 1955,
       1, 151, 1956,
       1, 152, 1957,
       1, 141, 1958,
       1, 159, 1959)

# Make Z into a matrix with 3 columns
Z <- matrix(Z, ncol = 3, byrow = TRUE)

# Calculate m1
m1 <- t(Z) %*% Z

# Calculate m2
m2 <- solve(m1)

# Calculate m3
m3 <- t(Z) %*% y

# Calculate estimator beta
beta <- m2 %*% m3

# Print estimator beta
cat("\nThe estimator is", beta)

# Compute mean of y
y.bar <- mean(y)

# Compute TSS
TSS <- sum( ( y - y.bar) ^ 2 )

# Compute predictions hat y
y.hat <- Z %*% beta

# Compute ESS
RSS <- sum( (y - y.hat) ^ 2 )

# Compute coefficient of determination R^2
R2 <- 1 - RSS / TSS

# Print R^2
cat("\nThe coefficient of determination R^2 is", R2)
```

<br>

The coefficient of determination
$$
R^2 = 0.8655401
$$
is quite close to $1$, showing that:

- The model fits data quite well
- *Productivity Index* and *Year* affect *Unemployment*
- *Unemployment* is almost completely explained by *Productivity Index* and *Year*  
(Because $R^2$ is almost $1$)




# Part 6: <br>Two sample t-test<br> as general regression {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## Two-sample t-test as general regression {.smaller}

- Two-sample t-test is special case of general linear regression

- This example is important for two reasons

  * Shows that multiple linear regression encapsulates simple models like the two-sample t-test
  * Shows that general linear regression leads to intuitive solutions in simple examples




## Two-sample t-test setting {.smaller}

- We have 2 populations $A$ and $B$ distributed like $N(\mu_A, \sigma_A^2)$ and $N(\mu_B, \sigma_B^2)$

- We have two samples
  * Sample of size $n_A$ from population $A$
  $$
  a = (a_1, \ldots, a_{n_A})
  $$
  * Sample of size $n_B$ from population $B$
  $$
  b = (b_1, \ldots, b_{n_B})
  $$

- Two-sample t-test compares means $\mu_A$ and $\mu_B$ by computing t-statistic
$$
t = \frac{ \overline{a} - \overline{b} }{\ese}
$$



## Setting up the regression analysis {.smaller}

- We want a regression model that can describe the sample means $\overline{a}$ and $\overline{b}$

- It is sufficient to consider a general linear model with $p = 2$, that is,

$$
\Expect[Y | Z_1 = z_1 , Z_2 = z_2 ] = \beta_1 z_1 + \beta_2 z_2
$$

- Define the data vector $y$ by concatenating $a$ and $b$

$$
y = 
\left(\begin{array}{c}
     y_1   \\
     \ldots \\
     y_{n_A} \\
     y_{n_A + 1}\\
     \ldots \\
     y_{n_A + n_B}
\end{array}\right)
= 
\left(\begin{array}{c}
     a_1   \\
     \ldots \\
     a_{n_A} \\
     b_{1}\\
     \ldots \\
     b_{n_B}
\end{array}\right)
$$







## Setting up the regression analysis {.smaller}

- The variable $Z_1$ is modelled as

$$
Z_1 = 1_{A}(i) := 
\begin{cases}
1 & \text{ if i-th observation is from population A} \\
0 & \text{ otherwise}
\end{cases}
$$


- Therefore, we have 

\begin{align*}
z_1 & = (\underbrace{1_A(i) , \ldots, 1_A(i)}_{n_A + n_B} ) \\[15pts]
    & = (\underbrace{1 , \ldots, 1}_{n_A}, 
         \underbrace{0 , \ldots, 0}_{n_B} )
\end{align*}


## Setting up the regression analysis {.smaller}

- Similarly, the variable $Z_2$ is modelled as

$$
Z_2 = 1_{B}(i) := 
\begin{cases}
1 & \text{ if i-th observation is from population B} \\
0 & \text{ otherwise}
\end{cases}
$$

- Therefore we have 

\begin{align*}
z_2 & = (\underbrace{1_B(i), \ldots, 1_B (i)}_{n_A + n_B} ) \\[15pts]
    & = (\underbrace{0 , \ldots, 0}_{n_A}, 
         \underbrace{1 , \ldots, 1}_{n_B} )
\end{align*}



## Setting up the regression analysis {.smaller}

- The general regression model is

$$
Y_i = \beta_1 \, 1_{A}(i) + \beta_2 \, 1_{B}(i) + \e_i
$$

- Variables $1_A$ and $1_B$ are called **dummy variables** (more on this later)

- The design matrix is therefore

$$
Z = (1_A | 1_B) =
\left(\begin{array}{cc}
     1_A(i) & 1_B (i)  \\
     \ldots & \ldots\\
     1_A(i) & 1_B (i) \\
\end{array}\right)  
 = 
\left(\begin{array}{cc}
     1 & 0  \\
     \ldots & \ldots\\
     1  & 0\\
     0 & 1 \\
     \ldots & \ldots\\
     0 & 1 
\end{array}\right)
$$





## Calculating $Z^T y$ {.smaller}

- We want to compute the regression estimator

$$
\hat \beta = (Z^T Z)^{-1} Z^T y
$$


- $Z$ has dimension $(n_A + n_B) \times 2$
- $Z^T$ has dimension $2 \times (n_A + n_B)$
- $y$ has dimension $(n_A + n_B) \times 1$
- $Z^T y$ therefore has dimension

$$
[ 2 \times (n_A + n_B)] \times [ (n_A + n_B) \times 1 ] 
= 2 \times 1
$$




## Calculating $Z^T y$ {.smaller}



\begin{align*}
Z^T y & = 
\left(\begin{array}{cc}
     1 & 0  \\
     \ldots & \ldots\\
     1 & 0\\
     0 & 1\\
     \ldots & \ldots\\
     0 & 1
\end{array}\right)^T
\left(\begin{array}{c}
     y_1  \\
     \ldots\\
     y_{n_A}\\
     y_{n_A+1}\\
     \ldots\\
     y_{n_A+n_B}
\end{array}\right) 
=
\left(\begin{array}{cccccc}
     1 & \ldots & 1 & 0 & \ldots & 0  \\
     0 & \ldots & 0 & 1 & \ldots & 1  \\
\end{array}\right)
\left(\begin{array}{c}
     y_1  \\
     \ldots\\
     y_{n_A}\\
     y_{n_A+1}\\
     \ldots\\
     y_{n_A+n_B}
\end{array}\right) \\[35pt]
& = 
\left(\begin{array}{c}
     y_1 \cdot 1 + \ldots + y_{n_A} \cdot 1 + 
     y_{n_A + 1} \cdot 0 + \ldots + y_{n_A + n_B} \cdot 0 \\
     y_1 \cdot 0 + \ldots + y_{n_A} \cdot 0 + 
     y_{n_A + 1} \cdot 1 + \ldots + y_{n_A + n_B} \cdot 1 
\end{array}\right) \\[20pt]
 & = 
\left(\begin{array}{c}
     y_1 + \ldots + y_{n_A} \\
     y_{n_A + 1} + \ldots + y_{n_A + n_B} 
\end{array}\right) 
 = 
\left(\begin{array}{c}
     a_1 + \ldots + a_{n_A} \\
     b_{1} + \ldots + b_{n_B} 
\end{array}\right) 
=
\left(\begin{array}{c}
     n_A \ \overline{a} \\
      n_B \ \overline{b}
\end{array}\right)
\end{align*}





## Calculating $Z^T Z$ {.smaller}

- $Z$ has dimension $(n_A + n_B) \times 2$
- $Z^T$ has dimension $2 \times (n_A + n_B)$
- $Z^T Z$ therefore has dimension

$$
[ 2 \times (n_A + n_B)] \times [ (n_A + n_B) \times 2 ] 
= 2 \times 2
$$


## Calculating $Z^T Z$ {.smaller}

\begin{align*}
Z^T Z & = 
\left(\begin{array}{cc}
     1 & 0  \\
     \ldots & \ldots\\
     1 & 0\\
     0 & 1\\
     \ldots & \ldots\\
     0 & 1
\end{array}\right)^T
\left(\begin{array}{cc}
     1 & 0  \\
     \ldots & \ldots\\
     1 & 0\\
     0 & 1\\
     \ldots & \ldots\\
     0 & 1
\end{array}\right) \\[35pt]
& = 
\left(\begin{array}{cccccc}
     1 & \ldots & 1 & 0 & \ldots & 0  \\
     0 & \ldots & 0 & 1 & \ldots & 1  \\
\end{array}\right)
\left(\begin{array}{cc}
     1 & 0  \\
     \ldots & \ldots\\
     1 & 0\\
     0 & 1\\
     \ldots & \ldots\\
     0 & 1
\end{array}\right)
= \left(\begin{array}{cc}
   n_A  & 0  \\
    0 & n_B
\end{array}\right)
\end{align*}


## Calculating $\hat \beta$ {.smaller}

- As $Z^T Z$ is diagonal, the inverse is

$$
(Z^T Z)^{-1} =
\left(\begin{array}{cc}
   n_A  & 0  \\
    0  & n_B
\end{array}\right)^{-1}
=\left(\begin{array}{cc}
    \frac{1}{n_A} & 0 \\
    0 & \frac{1}{n_B}
\end{array}\right)
$$


- The MLE is therefore

$$
\hat \beta = (Z^T Z)^{-1} Z^Ty = 
\left(\begin{array}{cc}
    \frac{1}{n_A} & 0 \\
    0 & \frac{1}{n_B}
\end{array}\right)
\left(\begin{array}{c}
     n_A \ \overline{a}  \\
      n_B \ \overline{b}
\end{array}\right)
=
\left(\begin{array}{c}
     \overline{a}  \\
     \overline{b} 
\end{array}\right)
$$


## Conclusions {.smaller}

- The MLE is

$$
\hat \beta = 
\left(\begin{array}{c}
     \overline{a}  \\
     \overline{b} 
\end{array}\right)
$$

- The regression function is hence

\begin{align*}
\Expect[Y | 1_A = z_1 , 1_B = z_2] & = \hat\beta_1 \ z_1 + \hat\beta_2 \ z_2 \\[20pt]
& = \overline{a} \ z_1 + \overline{b} \ z_2 
\end{align*}


## Conclusions {.smaller}

- This is a very intuitive result:
  * The expected value for $Y$ when data is observed from population $A$ is
  $$
  \Expect[Y | 1_A = 1 , 1_B = 0] = \overline{a}
  $$
  * The expected value for $Y$ when data is observed from population $B$ is
  $$
  \Expect[Y | 1_A = 0 , 1_B = 1] = \overline{b}
  $$


- Therefore, under this simple model
  * The best estimate for population A is the sample mean $\overline{a}$
  * The best estimate for population B is the sample mean $\overline{b}$